---
title: "Differential Expression Explorer"
format:
  live-dashboard:
    orientation: rows
    engine: knitr
    theme:
      light: 
        - sandstone
        - dashboard.scss
webr:
  packages:
    - dplyr
    - dbplyr
    - DBI
    - duckdb
    - ggplot2
    - ggrepel
    - svglite
    - jsonlite
    - plotly
  repos:
    - https://r-lib.r-universe.dev
---

::: {.content-hidden profile="local"}
<!-- Note: do not remove the following line. It is used when rendering. -->
REPLACEME
:::

::: {.content-visible profile="local"}
{{< include ../../_extensions/r-wasm/live/_knitr.qmd >}}
:::

Welcome to the **Differential Expression Explorer**, a serverless dashboard for interactive analysis of CosMx SMI differential expression results. This tool utilizes WebAssembly [(WebR)](https://docs.r-wasm.org/webr/latest/){target="_blank"} and [ObservableJS](https://quarto.org/docs/interactive/ojs/){target="_blank"} to run entirely in your web browser, allowing you to visualize **Volcano Plots**, **Marginal Means**, and **Spatial Mappings** without uploading data to a remote server. You can explore existing DE results that 
we've added or load your own results. Advanced users should check out the `Advanced`
tab if they want to customize their volcano plot even further. 

::: {.callout-warning appearance="simple"}
**Beta Notice:** This dashboard is currently in an **experimental / beta testing phase**. Features, performance, and data formats are subject to change. When you first load the site, you will see temporary `
OJS Runtime Error` messages displayed until data are initialized. Please report any issues you encounter.
:::

# {.sidebar width="500px"}

:::{.card .hidden}

```{ojs}
//| echo: false
//| output: false

duck_db_client = {
  const duckdb = await import("https://cdn.jsdelivr.net/npm/@duckdb/duckdb-wasm@1.28.0/+esm");
  
  console.log("🔧 Initializing DuckDB-Wasm Client...");
  
  const JSDELIVR_BUNDLES = duckdb.getJsDelivrBundles();
  const bundle = await duckdb.selectBundle(JSDELIVR_BUNDLES);
  
  const worker_url = URL.createObjectURL(
    new Blob([`importScripts("${bundle.mainWorker}");`], { type: "text/javascript" })
  );

  const worker = new Worker(worker_url);
  const logger = new duckdb.ConsoleLogger();
  const db = new duckdb.AsyncDuckDB(logger, worker);
  
  await db.instantiate(bundle.mainModule, bundle.pthreadWorker);
  
  const conn = await db.connect();
  conn.instance = db; // Attach the engine to the connection
  console.log("✅ DuckDB-Wasm Ready (Engine + Connection)");
  return conn;
}
```

```{webr}
#| echo: false
#| output: false
#| autorun: true
#| input: 
#|   - selected_study

if(exists("con") && dbIsValid(con)){
  cat("Closing previous connection!")
  dbDisconnect(con)
} 
cat("\nCreating new connection.")
con <- dbConnect(duckdb())
```

```{webr}
#| echo: false
#| autorun: true

# function to does a quick verification of user data
verify <- function(vfs_dir){
  valid <- TRUE
  files <- list.files(vfs_dir)
  required_files <- c(
    "cell_metadata.parquet", "emmeans.parquet",
    "pairwise.parquet", "study_header.parquet"
  )
  if(!all(required_files %in% files)){
    cat(paste0("The zip file must contain the following: ", paste0(
      required_files, collapse=", "), "."))
    valid <- FALSE
  }
  return(valid)
}
```

```{ojs}
//| echo: false
//| output: false

sparkbar = (max) => {
  return (x) => {
    // 1. Create container
    const div = d3.create("div")
      .style("display", "flex")
      .style("align-items", "center")
      .style("width", "100%"); // Let it fill the cell

    const color = x >= 0 ? "#E41A1C" : "#377EB8"; 
    const widthPercent = Math.min((Math.abs(x) / max) * 100, 100);

    div.append("span")
       .text(x.toFixed(2))
       .style("font-size", "0.85em")
       .style("color", "#666")
       .style("width", "35px")         // Fixed width for alignment
       .style("text-align", "right")   // Align numbers to the right
       .style("margin-right", "8px");  // Space between number and bar

    div.append("div")
      .style("width", `${widthPercent}%`)
      .style("height", "12px")
      .style("background-color", color)
      .style("border-radius", "2px");

    return div.node();
  }
}
```

```{ojs}
//| echo: false
//| output: false

d3 = require("d3@7")
Plotly = require("https://cdn.plot.ly/plotly-2.24.1.min.js")
manifest_raw = FileAttachment("https://raw.githubusercontent.com/Nanostring-Biostats/ScratchSpaceDatasets/refs/heads/main/browser_bytes/manifest.tsv").tsv({typed: true})

mutable available_studies = {
  if (!manifest_raw) return [];

  return manifest_raw.map(d => ({
    id: d.id,
    name: d.name,
    description: d.description,
    
    // Direct mapping of your TSV columns
    url_header:   d.url_header,
    url_pairwise: d.url_pairwise,
    url_meta:     d.url_meta,
    url_emmeans:  d.url_emmeans,
    url_overlap: d.url_overlap,

    type: 'remote',
    valid: true
  }));
}
```
:::


```{ojs}
//| echo: false
//| output: true

viewof selected_study = Inputs.select(available_studies, {
  label: "Select Study:",
  format: d => d.name,
  value: latest_upload || available_studies[0] 
})
```

```{ojs}
//| echo: false

{
  const data = study_data_payload;
  const contrast = contrast_selection; 
  const volcano = current_volcano_data || [];

  if (!data || !data.header_data || data.header_data.length === 0) {
    return html`<div style="font-size: 0.9em; color: #888; font-style: italic;">⏳ Waiting for study data...</div>`;
  }

  const row = data.header_data[0];
  const pretty_contrast = contrast ? contrast.replace(" / ", " vs. ") : "None selected";
  const count = volcano.length ? volcano.length.toLocaleString() : "0";

  return html`
  <div style="font-size: 0.9em; color: #444;">
    
    <h5 style="text-transform: uppercase; color: #888; font-size: 0.75em; margin-top: 20px; margin-bottom: 4px; border-top: 1px solid #ddd; padding-top: 12px;">
      Study Description
    </h5>
    <p style="margin-bottom: 16px; font-weight: 500;">${row.Description || "No description available."}</p>
    
    <h5 style="text-transform: uppercase; color: #888; font-size: 0.75em; margin-top: 15px; margin-bottom: 8px; border-top: 1px solid #ddd; padding-top: 12px;">
      Model Formula
    </h5>
    <div style="background: #f4f4f4; padding: 10px; border-radius: 6px; font-family: monospace; color: #d63384; font-size: 0.9em; margin-bottom: 20px;">
      ${row.Formala || "NA"}
    </div>
    
    <div style="display: flex; gap: 20px; border-top: 1px solid #ddd; padding-top: 15px;">
      <div style="flex: 1;">
        <h5 style="text-transform: uppercase; color: #888; font-size: 0.75em; margin-bottom: 4px;">Current Contrast</h5>
        <div style="font-weight: 600; color: #007bff;">${pretty_contrast}</div>
      </div>
      <div style="flex: 0 0 100px; border-left: 1px solid #ddd; padding-left: 15px;">
        <h5 style="text-transform: uppercase; color: #888; font-size: 0.75em; margin-bottom: 4px;">Total Targets</h5>
        <div style="font-weight: bold; font-size: 1.1em;">${count}</div>
      </div>
    </div>
    
    <h5 style="text-transform: uppercase; color: #888; font-size: 0.75em; margin-top: 20px; margin-bottom: 4px; border-top: 1px solid #ddd; padding-top: 12px;">
    </h5>
  </div>
  `;
}
```

<h5 style="text-transform: uppercase; color: #888; font-size: 0.75em; margin-top: 20px; margin-bottom: 8px; border-top: 1px solid #ddd; padding-top: 12px;">
  Data Generation Options
</h5>

You can use your own results in this dashboard! Whether your DE results come 
from AtoMx SIP's DE Module or the open-sourced smiDE package, we can show you
how to load your results.

* **AtoMx SIP**: ✅ [Convert data in browser](#sec-convert-atomx)
* **smiDE (R Package)**: ✅ [View instructions](#sec-convert-smide)

<div style="background: #fdfdfe; border: 1px solid #eee; padding: 10px; border-radius: 6px; margin-top: 10px;">
  <p style="margin-bottom: 0; font-size: 0.85em; color: #666; font-style: italic;">
    Note: Both methods generate a dashboard-ready .zip file.
  </p>
</div>

```{ojs}
//| echo: false
//| output: true

viewof new_zip_file = Inputs.file({
  label: "Import the formatted results (.zip):",
  accept: ".zip",
  multiple: false
})
```

:::{.card .hidden}

```{ojs}
//| echo: false
//| output: false

// 1. Declare the container as mutable
mutable new_zip_data = null;

// 2. This "watcher" cell handles manual file uploads
{
  if (new_zip_file) {
    const buf = await new_zip_file.arrayBuffer();
    mutable new_zip_data = {
      id: Date.now(),
      filename: new_zip_file.name,
      data: new Uint8Array(buf)
    };
  }
}
```

```{webr}
#| echo: false
#| autorun: true
#| output: false
#| input: 
#|   - new_zip_data
#| define: 
#|   - processed_new_study

processed_new_study <- NULL

if(length(new_zip_data) > 1){
  cat(paste0("Processing new user data. Timestamp: ", Sys.time()))

  zip_file_name <- as.character(new_zip_data$filename)
  target_dir <- as.character(new_zip_data$id)

  writeBin(as.raw(new_zip_data$data), zip_file_name)
  if(dir.exists(target_dir)) unlink(target_dir, recursive=TRUE)
  utils::unzip(zip_file_name, exdir = target_dir, junkpaths = TRUE)
  if(file.exists(zip_file_name)) unlink(zip_file_name, recursive=TRUE)

  isValid <- verify(target_dir)
  
  if(!isValid){
    cat("\n❌ Validation Failed.")
  } else {
    header_path <- file.path(target_dir, "study_header.parquet")
    study_name <- paste0("Custom: ", zip_file_name) 
    header_df <- data.frame(Name = study_name, Description = "User Data", Formala = "NA") # Fallback

    tryCatch({
       con_peek <- dbConnect(duckdb())
       df <- dbGetQuery(con_peek, paste0("SELECT * FROM '", header_path, "' LIMIT 1"))
       
       if(nrow(df) > 0) {
         header_df <- df
         if("Name" %in% names(df)) study_name <- as.character(df$Name[1])
       }
       dbDisconnect(con_peek, shutdown=TRUE)
    }, error = function(e) {
       if(exists("con_peek") && dbIsValid(con_peek)) dbDisconnect(con_peek, shutdown=TRUE)
       cat("Warning: Could not read header file. Using defaults.\n")
    })
    
    cat(paste0("\n✅ Validated: ", study_name))

    new_study_list <- list(
      id = target_dir,
      name = study_name,
      description = "User Uploaded Data", 
      
      header_data = header_df, 
      
      url_header   = header_path,
      url_pairwise = file.path(target_dir, "pairwise.parquet"),
      url_meta     = file.path(target_dir, "cell_metadata.parquet"),
      url_emmeans  = file.path(target_dir, "emmeans.parquet"),
      url_overlap = file.path(target_dir, "overlap.parquet"),
      type = "local",
      valid = TRUE
    )
    
    processed_new_study <- as.character(jsonlite::toJSON(new_study_list, auto_unbox = TRUE))
  }
  
} else {
  cat(paste0("new_zip_data is NA. Timestamp: ", Sys.time()))
}
```

```{ojs}
//| echo: false
//| output: false

mutable latest_upload = null
```

```{ojs}
//| echo: false
//| output: false

{
  if (processed_new_study) {
    const new_record = JSON.parse(processed_new_study);
    
    if (new_record && new_record.valid) {
      const exists = available_studies.find(d => d.id === new_record.id);
      
      if (!exists) {
         // 1. Update the list
         mutable available_studies.push(new_record);
         mutable available_studies = mutable available_studies;
         
         mutable latest_upload = new_record; 
      }
      console.log("✅ Loaded:", new_record.name);
    }
  }
}
```

```{ojs}
//| echo: false
//| output: false

study_data_payload = {
  const selection = selected_study;
  if (!selection) return null;

  console.log("🔄 Processing selection:", selection.name);

  if (selection.type === "local") {
    console.log("👉 Mode: Local");
    return {
      mode: "local",
      ...selection
    };
  }

  if (selection.type === "remote") {
    console.log("👉 Mode: Remote (JS Bridge)");
    
    try {
      const conn = await duck_db_client;
      
      const header_table = await conn.query(`SELECT * FROM '${selection.url_header}'`);
      
      const header_json = header_table.toArray().map(row => row.toJSON());
      
      console.log("✅ Header received:", header_json);

      return {
        mode: "remote_js_bridge",
        id: selection.id,
        name: selection.name,
        header_data: header_json,
        url_pairwise: selection.url_pairwise,
        url_meta: selection.url_meta,
        url_emmeans: selection.url_emmeans,
        url_overlap: selection.url_overlap
      };
      
    } catch (e) {
      console.error("❌ JS Bridge Error:", e);
      return null;
    }
  }
}
```

```{webr}
#| echo: false
#| autorun: true
#| output: false
#| input: 
#|   - study_data_payload
#| define: 
#|   - db_ready
#|   - r_status_proxy

db_ready <- FALSE
r_status_proxy <- "Waiting for data..."
print(con)

if (!is.null(study_data_payload)) {
  
  cat(paste0("R: Received payload mode: ", study_data_payload$mode, "\n"))

  if (study_data_payload$mode == "remote_js_bridge") {
    
    if (!is.null(study_data_payload$header_data)) {
       header_json_str <- jsonlite::toJSON(study_data_payload$header_data)
       header_df <- jsonlite::fromJSON(header_json_str)

       dbWriteTable(con, "study_meta", header_df, overwrite = TRUE)
       cat("R: Header table written successfully.\n")
    }
    
  } else {

    if (!is.na(study_data_payload$url_header)) {
       dbExecute(con, paste0("CREATE OR REPLACE VIEW study_meta AS SELECT * FROM '", study_data_payload$url_header, "'"))
    }

    if (!is.na(study_data_payload$url_pairwise)) {
       dbExecute(con, paste0("CREATE OR REPLACE VIEW de_results AS SELECT * FROM '", study_data_payload$url_pairwise, "'"))
    }

    if (!is.na(study_data_payload$url_meta)) {
       dbExecute(con, paste0("CREATE OR REPLACE VIEW cell_meta AS SELECT * FROM '", study_data_payload$url_meta, "'"))
    }

    if (!is.na(study_data_payload$url_emmeans)) {
       dbExecute(con, paste0("CREATE OR REPLACE VIEW emmeans_data AS SELECT * FROM '", study_data_payload$url_emmeans, "'"))
    }
    # Overlap (furture feature)
    #if (!is.na(study_data_payload$url_overlap)) {
    #   dbExecute(con, paste0("CREATE OR REPLACE VIEW overlap_data AS SELECT * FROM '", study_data_payload$url_overlap, "'"))
    #}
      }
  
  r_status_proxy <- paste("Active:", study_data_payload$name)
  db_ready <- TRUE
}
```

```{ojs}
//| echo: false
//| output: false

contrasts_from_js = {
  const payload = study_data_payload;
  // If local, let webR/R handle the contrast list to avoid path errors
  if (!payload || payload.type === "local") return null;

  if (payload.mode === "remote_js_bridge") {
    try {
      const conn = await duck_db_client;
      const table = payload.isAtoMx ? 'pairwise_harmonized' : `"${payload.url_pairwise}"`;
      
      const query = `SELECT DISTINCT "contrast" FROM ${table} ORDER BY "contrast"`;
      const result = await conn.query(query);
      return result.toArray().map(row => row.contrast).filter(d => !d.includes("otherct_expr"));
    } catch (e) {
      console.error("❌ Contrast Fetch Error:", e);
      return []; 
    }
  }
}
```

```{ojs}
//| echo: false
//| output: false

terms_from_js = {
  const payload = study_data_payload;
  // Guard against local studies to allow R-based term population
  if (!payload || payload.type === "local") return null;

  if (payload.mode === "remote_js_bridge") {
    try {
      const conn = await duck_db_client;
      const table = payload.isAtoMx ? 'pairwise_harmonized' : `"${payload.url_pairwise}"`;
      
      const query = `SELECT DISTINCT "term" FROM ${table} ORDER BY "term"`;
      const result = await conn.query(query);
      return result.toArray().map(row => row.term);
    } catch (e) {
      console.error("❌ Terms Fetch Error:", e);
      return [];
    }
  }
}
```

```{ojs}
//| echo: false
//| output: false

slides_from_js = {
  const payload = study_data_payload;
  // If local, return null to let R query the local cell_meta table
  if (!payload || payload.type === "local") return null;

  if (payload.mode === "remote_js_bridge") {
    if (payload.isAtoMx) return [1]; 

    try {
      const conn = await duck_db_client;
      const query = `SELECT DISTINCT "slide_id_numeric" FROM "${payload.url_meta}" ORDER BY "slide_id_numeric"`;
      const result = await conn.query(query);
      const list = result.toArray().map(row => row.slide_id_numeric);
      return list.length > 0 ? list : [1];
    } catch (e) {
      return [1];
    }
  }
}
```

```{webr}
#| echo: false
#| autorun: true
#| output: false
#| input: 
#|   - study_data_payload
#|   - contrasts_from_js
#|   - terms_from_js
#|   - slides_from_js
#| define: 
#|   - available_contrasts
#|   - available_terms
#|   - available_slides

# print(Sys.time())
# print(contrasts_from_js)
# print(terms_from_js)
# print(slides_from_js)

available_contrasts <- if(exists("available_contrasts")) available_contrasts else NULL
available_terms <- if(exists("available_terms")) available_terms else NULL
available_slides <- if(exists("available_slides")) available_slides else 1

if (!is.null(study_data_payload)) {
  # REMOTE MODE
  if (study_data_payload$mode == "remote_js_bridge") {
    if (!is.null(contrasts_from_js)) available_contrasts <- unlist(contrasts_from_js)
    if (!is.null(terms_from_js)) available_terms <- unlist(terms_from_js)
    if (!is.null(slides_from_js)) available_slides <- unlist(slides_from_js)
    
  # LOCAL MODE
  } else if (dbExistsTable(con, "de_results")) {
    available_contrasts <- dbGetQuery(con, "SELECT DISTINCT contrast FROM de_results")$contrast
    available_contrasts <- available_contrasts[!grepl("otherct_expr", available_contrasts)]
    available_terms <- dbGetQuery(con, "SELECT DISTINCT term FROM de_results")$term
    
    if (dbExistsTable(con, "cell_meta")) {
      available_slides <- dbGetQuery(con, "SELECT DISTINCT slide_id_numeric FROM cell_meta")$slide_id_numeric
    }
  }
}
```

:::

:::{.card}

<br>

For best results, use Chrome, Edge, or Safari. This is optimized for desktop and 
tablet view.

Tool Version: 1.0

Tool Type: Serverless (OJS + webr)

Source Code Available:     ✅ Yes



:::





# Volcano Plots

## Row {height=100%}


### Column {width=60%}


#### Row {height=50%}

##### Column {width=50%}

###### Row {height=100%}

::: {.card title="SUMMARY STATISTICS"}

```{webr}
#| echo: false
#| autorun: true
#| input: 
#|   - vplot_stats
#| results: asis

if (inherits(vplot_stats, "data.frame")) {
  knitr::kable(vplot_stats, format = "html", 
               col.names = c("", ""), # Hide headers if self-explanatory
               table.attr = "style='width:100%; font-size: 0.9em; border-collapse: collapse;'")
} else {
  cat("<span style='color:#ccc'>Waiting for data...</span>")
}
```

:::

##### Column {width=50%}

```{ojs}
//| title: ADJUST IMAGE AESTHETICS
//| echo: false
//| scrolling: true

viewof aesthetics_raw = Inputs.form({
  volcano_fc_range:      Inputs.range([0, 10], {step: 0.001, label: "Abs. log2FC", value: 1}),
  volcano_pval_threshold: Inputs.range([0, 1], {step: 0.00001, label: "P-value threshold", value: 0.05}),
  volcano_pt_size:       Inputs.range([0.1, 5], {step: 0.1, label: "Pt. Size"}),
  volcano_alpha:         Inputs.range([0, 1], {step: 0.01, label: "Pt. Alpha", value: 1.0}),
  
  volcano_pt_color:      Inputs.color({label: "Pt. Color", value: "#3C3F40"}),
  volcano_sig_color:     Inputs.color({label: "Significant targets", value: "#b44697"}),
  
  volcano_high_fc_label: Inputs.range([0, 10], {step: 1, label: "Label high FC", value: 3}),
  volcano_low_fc_label:  Inputs.range([0, 10], {step: 1, label: "Label low FC", value: 3}),
  volcano_pval_label:    Inputs.range([0, 10], {step: 1, label: "Label Sig.", value: 3}),
  volcano_include_selected_genes: Inputs.radio(["Yes", "No"], {label: "Label selected?", value: "Yes"})
})

aesthetics_debounced = {
  const raw = aesthetics_raw;
  return new Promise(resolve => {
    const timer = setTimeout(() => resolve(raw), 600); 
    invalidation.then(() => clearTimeout(timer));
  });
}


aesthetics = aesthetics_debounced

volcano_fc_range               = aesthetics.volcano_fc_range
volcano_pval_threshold         = aesthetics.volcano_pval_threshold
volcano_pt_size                = aesthetics.volcano_pt_size
volcano_alpha                  = aesthetics.volcano_alpha
volcano_pt_color               = aesthetics.volcano_pt_color
volcano_sig_color              = aesthetics.volcano_sig_color
volcano_high_fc_label          = aesthetics.volcano_high_fc_label
volcano_low_fc_label           = aesthetics.volcano_low_fc_label
volcano_pval_label             = aesthetics.volcano_pval_label
volcano_include_selected_genes = aesthetics.volcano_include_selected_genes

```


#### Row {height=50%}

##### Column {width=100%}

:::{.card .hidden}

```{ojs}
//| echo: false
//| output: false

volcano_data_js = {
  const payload = study_data_payload;
  const contrast = contrast_selection;
  const term = term_selection;
  
  // 1. DONT PROCEED if essential selections are missing
  if (!payload || !contrast || !term || payload.mode === "local") return null;

  if (payload.mode === "remote_js_bridge") {
    try {
      const conn = await duck_db_client;
      const tableSource = payload.isAtoMx ? `pairwise_harmonized` : `"${payload.url_pairwise}"`;

      // 2. CHECK IF VIEW EXISTS (For AtoMx)
      if (payload.isAtoMx) {
         const viewCheck = await conn.query(`SELECT count(*) as n FROM ${tableSource}`);
         if (viewCheck.toArray()[0].n === 0) return null; // Wait for data
      }

      const query = `
        SELECT term, contrast, target, log2fc, "p.value", ncells_1, ncells_2
        FROM ${tableSource}
        WHERE contrast = '${contrast}' AND term = '${term}'
      `;
      
      const result = await conn.query(query);
      const data = result.toArray().map(row => row.toJSON());
      
      // 3. ONLY RETURN IF WE HAVE DATA
      if (data.length === 0) return null; 

      console.log(`✅ Pairwise data fetched from ${tableSource} (${data.length} rows)`);
      return data;
      
    } catch (e) {
      console.error("❌ Failed to fetch pairwise data:", e);
      return null; // Return null so webR doesn't try to process an error
    }
  }
}
```

```{webr}
#| echo: false
#| autorun: true
#| output: false
#| input: 
#|   - study_data_payload
#|   - volcano_data_js
#|   - contrast_selection
#|   - term_selection
#| define: 
#|   - current_volcano_data

current_volcano_data <- data.frame('term'='unknown', 
                                   'contrast' = 'B / A',
                                   'target' = 'NA',
                                   'log2fc' = 0,
                                   'p.value' = 1,
                                   'ncells_1' = 0,
                                   'ncells_2' = 0)

if (!is.null(study_data_payload) && !is.null(contrast_selection)) {

  if (study_data_payload$mode == "remote_js_bridge") {
    if (!is.null(volcano_data_js)) {
       json_str <- jsonlite::toJSON(volcano_data_js)
       current_volcano_data <- jsonlite::fromJSON(json_str)
       
       cat(paste0("R: Received ", nrow(current_volcano_data), " rows from Remote.\n"))
    }
    
  } else {
    
    cat(paste0("R: Querying Local DB for: ", contrast_selection, "\n"))

    if (dbExistsTable(con, "de_results")) {
        query <- sprintf(
          "SELECT term, contrast, target, log2fc, \"p.value\", ncells_1, ncells_2 
           FROM de_results 
           WHERE contrast = '%s' AND term = '%s'", 
          contrast_selection, 
          term_selection
        )

       current_volcano_data <- dbGetQuery(con, query)
       cat(paste0("R: Loaded ", nrow(current_volcano_data), " rows from Local.\n"))
    }
  }
  
  current_volcano_data <- current_volcano_data %>% mutate(log2fc = log2(fold_change), padj = p.adjust(p.value, method="BH")) %>%
    select(target, log2fc, p.value, padj, ncells_1, ncells_2, term, contrast)
}
```

:::

```{ojs}
//| echo: false
//| content: card-toolbar
//| input:  current_volcano_data
viewof pairwise_search = Inputs.search(current_volcano_data, {placeholder: "Search Results"})

```

::: {.card title="PAIRWISE DETAILS"}

```{ojs}
//| echo: false
//| output: true

viewof pairwise_table = Inputs.table(pairwise_search, {
  multiple: true,
  rows: 20,
  format: {
    log2fc: sparkbar(d3.max(pairwise_search, d => d.log2fc))
  }})

```

<!-- ```{ojs} -->
<!-- selected_genes -->
<!-- ``` -->


```{ojs}
//| echo: false
//| output: false

// captures any selected genes (up to 30)
selected_genes = {
  if (!current_volcano_data || !pairwise_table) return ['notagene'];
  if (pairwise_table.length === current_volcano_data) {
    return ['notagene'];
  }
  if (pairwise_table.length > 30) {
    return ['notagene'];
  }
  return pairwise_table.map(d => d.target);
}

selectionWarning = {
  // 1. Wait for data
  if (!pairwise_table || !current_volcano_data) return `Select between 1 and 30 genes to label.`;

  const count = pairwise_table.length;
  const total = current_volcano_data.length;
  const isSelectAll = (count === total);

  if(isSelectAll){
    return `Select up to 30 genes to label in the volcano plot.`;
  }
  if (count > 30) {
    return `⚠️ Too many targets selected (${count}). Limit is 30.`;
  }

  return `Currently selected: ${count} gene${count === 1 ? '' : 's'}.`;
}
```

${selectionWarning}

:::

### Column {width=40%}



```{ojs}
//| echo: false
//| content: card-toolbar

// must occur before volcano plot card
viewof term_selection = {
  if (!available_terms || available_terms.length === 0) return md`_Loading terms..._`;

  return Inputs.select(available_terms, {
    label: "Available Terms:",
    multiple: false,
    size: 3
  });
}
viewof contrast_selection = {
  if (!available_contrasts || available_contrasts.length === 0) return md`_Loading contrasts..._`;

  return Inputs.select(available_contrasts, {
    label: "Available Contrasts:",
    format: d => d.replace(" / ", " vs. "),
    multiple: false,
    size: 3
  });
}

```


::: {.card title="VOLCANO PLOT"}


```{webr}
#| warning: false
#| echo: false
#| autorun: true
#| input: 
#|    - volcano_plot_emit
#|    - argsVolcano
#|    - current_volcano_data
#| define:
#|    - vplot_stats
#|    - vplot_plot
#|    - vplot_labels

if(!inherits(current_volcano_data, "data.frame") && is.na(current_volcano_data)){
    cat("Waiting on contrast selection.")
  vplot_stats <- ""
  vplot_labels <- c()
} else {
  argsCombined <- c(list(df=current_volcano_data %>% 
                           select(contrast, target, 
                                  log2fc, p.value, any_of(c("ncells_1", "ncells_2")))), argsVolcano)
  vplot_result <- rlang::exec("plotVolcano", !!!argsCombined)
  vplot_stats <- vplot_result[[2]]
  vplot_plot <- vplot_result[[1]]
  vplot_labels <- vplot_result[[3]]
  vplot_plot
}

# print(Sys.time())

```

<!-- Save Image -->

```{ojs}
//| echo: false
viewof save_settings = Inputs.form({
  width:  Inputs.number({label: "Width", value: 6, step: 0.5}),
  height: Inputs.number({label: "Height", value: 4, step: 0.5}),
  units:  Inputs.select(["in", "cm", "mm", "px"], {label: "Units", value: "in"}),
  format: Inputs.select(["png", "svg"], {label: "Format", value: "png"})
})
viewof save_btn = Inputs.button("Download Plot", {value: 0, color: "#2C3E50"})
save_request = {
  save_btn; 
  return save_settings;
}
```

```{webr}
#| echo: false
#| autorun: true
#| input: 
#|    - save_btn
#|    - save_request
#| define: 
#|   - downloaded_file

if(save_btn>0){
  cat('saving')
  fname <- paste0("volcano_plot.", save_request$format)
  ggsave(
    filename = fname, 
    plot = vplot_plot, 
    device = save_request$format,
    width = as.numeric(save_request$width),
    height = as.numeric(save_request$height),
    units = save_request$units
  )
  if(file.exists(fname)){
    # cat(paste0("file created: ", fname))
    file_bytes <- readBin(fname, "raw", n = file.info(fname)$size)
    downloaded_file <- list(
      data = file_bytes,
      filename = fname,
      mime = ifelse(save_request$format == "svg", "image/svg+xml", "image/png"),
      ts = as.character(Sys.time()) # Timestamp to force OJS update
    )
  }
}

```

```{ojs}
//| echo: false
//| output: false
{
  if (downloaded_file) {
    const blob = new Blob([new Uint8Array(downloaded_file.data)], { type: downloaded_file.mime });
    
    const url = URL.createObjectURL(blob);
    const a = document.createElement("a");
    a.href = url;
    a.download = downloaded_file.filename;
    document.body.appendChild(a);
    a.click();
    
    document.body.removeChild(a);
    URL.revokeObjectURL(url);
  }
}
```


:::

# Marginal Means

## Row {height=40%}

### Column {width=50%}

:::: {.card title="Description"}

This plot shows the estimated mean expression of a gene (_i.e._, marginal means) between 
the pairwise contrast that is currently selected. This can help you gauge the magnitude 
of the difference between the two groups after accounting for any other variables in the model.

Like the volcano plot, the x-axis shows the log2 fold change. The y-axis shows marginal mean response (log2 transformed).
Each gene is represented by up to two points (one for each level of the contrast) with error bars. When "vs rest" or "vs all" is
selected, only a single level is shown.

```{ojs}
//| echo: false
//| output: false

available_levels = {
  const payload = study_data_payload;
  // Prevents OJS from crashing on local VFS paths
  if (!payload || payload.type === "local") return null;
  
  if (payload.mode === "remote_js_bridge") {
    try {
      const conn = await duck_db_client;
      const query = `SELECT DISTINCT level FROM '${payload.url_emmeans}' ORDER BY level`;
      const result = await conn.query(query);
      return result.toArray().map(row => row.level).filter(t => !t.startsWith("otherct_"));
    } catch (e) {
      console.error("❌ Available Levels Fetch Error:", e);
      return [];
    }
  }
}
```

```{webr}
#| echo: false
#| autorun: true
#| input: 
#|    - study_data_payload
#|    - available_levels
#| define: 
#|    - r_available_levels

r_available_levels <- NULL
if (!is.null(study_data_payload)) {
  if (study_data_payload$mode == "remote_js_bridge") {
    r_available_levels <- available_levels
  } else {
    if (dbExistsTable(con, "emmeans_data")) {
        df <- dbGetQuery(con, "SELECT DISTINCT level FROM emmeans_data ORDER BY level")
        r_available_levels <- df$level[!grepl("^otherct_", df$level)]
    }
  }
}
```
::::

### Column {width=50%}

```{ojs}
//| echo: false
//| content: card-toolbar

viewof term_copy = {
  
  if (!available_terms || available_terms.length === 0) return md`_Loading..._`;

  const mirror_input = Inputs.select(available_terms, {
    label: "Available Terms:",
    multiple: false,
    size: 3
  });

  return Inputs.bind(mirror_input, viewof term_selection);
}

viewof contrast_copy = {

  if (!available_contrasts || available_contrasts.length === 0) return md`_Loading..._`;

  const mirror_input = Inputs.select(available_contrasts, {
    label: "Available Contrasts:",
    format: d => d.replace(" / ", " vs. "),
    multiple: false,
    size: 3
  });

  return Inputs.bind(mirror_input, viewof contrast_selection);
}
```

:::: {.card title="Controls"}

```{ojs}
//| echo: false

viewof means_raw = Inputs.form({
  means_highlight: Inputs.text({label: "Highlight Target:", placeholder: "e.g. A1BG,A1CF"}),
  means_show_labeled:       Inputs.toggle({label: "Also highlight Volcano-labeled Genes", value: true}),
  means_alpha: Inputs.range([0.1, 1], {value: 0.8, step: 0.1, label: "Dot Transparency"}),
  means_color_1: Inputs.color({label: "Level 1 Color", value: "#e15759"}),
  means_color_2: Inputs.color({label: "Level 2 Color", value: "#4e79a7"})
})

means_debounced = {
  const raw = means_raw;
  return new Promise(resolve => {
    const timer = setTimeout(() => resolve(raw), 600); 
    invalidation.then(() => clearTimeout(timer));
  });
}

means_highlight       = means_debounced.means_highlight
means_show_labeled    = means_debounced.means_show_labeled
means_alpha           = means_debounced.means_alpha
means_color_1         = means_debounced.means_color_1
means_color_2         = means_debounced.means_color_2

combined_selection = {

  const manual = means_highlight 
    ? means_highlight.toUpperCase().split(',').map(s => s.trim()).filter(s => s.length > 0) 
    : [];
    
  let from_volcano = [];
  if (means_show_labeled && typeof vplot_labels !== "undefined") {
      if(Array.isArray(vplot_labels)) {
        from_volcano = vplot_labels;
      } else if (typeof vplot_labels === 'string') {
        from_volcano = [vplot_labels];
      }
  }
  
  return Array.from(new Set([...manual, ...from_volcano]));
}
means_level = {
  const contrast = contrast_selection;
  if (!contrast) return [];

  return contrast.split(/\/|vs\./).map(s => s.trim());
}
```

::::

## Row {height=60%}

### Column {width=50%}

:::: {.card title='Marginal Means Plot'}

```{ojs}
//| echo: false
//| input: means_search

{

  if (!means_search || means_search.length === 0) {
    yield html`<div style="height:400px; display:flex; align-items:center; justify-content:center; background:#f9f9f9; border:1px border-radius:8px;">
      <p style="color:#888;">Select levels in the sidebar to visualize...</p>
    </div>`;
  } else {

    const colorScale = d3.schemeTableau10;
    
    let table_targets = [];
    
    if (typeof means_table !== "undefined" && means_table && means_search) {
        if (means_table.length !== means_search.length) {
          table_targets = means_table.map(d => d.target);
        }
    }
    
    const manual_targets = combined_selection || [];
    
    const highlight_list = Array.from(new Set([...table_targets, ...manual_targets]));
    
    const targetGroups = d3.groups(means_search, d => d.target);
    
    const levels = [...new Set(means_search.map(d => d.level))].sort();

    const traces = [];

    const alpha_scaler = means_alpha;
    const line_segments_high = { x: [], y: [], hoverinfo: 'skip', showlegend: false, mode: 'lines', line: {color: '#888', width: 1.5}, opacity: 0.8 * alpha_scaler, type: 'scatter' }; 
    const line_segments_med  = { x: [], y: [], hoverinfo: 'skip', showlegend: false, mode: 'lines', line: {color: '#888', width: 1}, opacity: 0.4 * alpha_scaler, type: 'scatter' };
    const line_segments_low  = { x: [], y: [], hoverinfo: 'skip', showlegend: false, mode: 'lines', line: {color: '#ccc', width: 1}, opacity: 0.3 * alpha_scaler, type: 'scatter' };

    targetGroups.forEach(([target, values]) => {
      if(values.length < 2) return;
      
      const v1 = values[0];
      const x = v1.log2fc;
      if(x == null || isNaN(x)) return;

      // Transform Response to Log2 for Y-axis
      const ys = values.map(d => Math.log2(d.response)); 
      
      const min_y = Math.min(...ys);
      const max_y = Math.max(...ys);
      
      const val_p = v1["p.value"];
      const p = val_p != null ? val_p : 1;
      
      let seg = line_segments_low;
      if(p < 0.001) seg = line_segments_high;
      else if(p < 0.05) seg = line_segments_med;
      
      seg.x.push(x, x, null);
      seg.y.push(min_y, max_y, null);
    });

    traces.push(line_segments_low, line_segments_med, line_segments_high);

    levels.forEach((lvl, i) => {
        const color = (i === 0) ? means_color_1 : means_color_2; 
        
        const lvl_trace = {
          x: [], y: [], text: [], ids: [],
          mode: 'markers',
          type: 'scatter',
          name: lvl, 
          marker: { color: color, size: 6, opacity: means_alpha },
          error_y: {
            type: 'data', symmetric: false, array: [], arrayminus: [],
            visible: true, color: color, thickness: 1, width: 2, opacity: 0.6
          }
        };

        const lvlData = means_search.filter(d => d.level === lvl);
        
        lvlData.forEach(d => {
           if (d.log2fc == null || isNaN(d.log2fc)) return;
           
           // Log2 Transformation
           const y_log = Math.log2(d.response);
           const ucl_log = Math.log2(d.UCL);
           const lcl_log = Math.log2(d.LCL);

           lvl_trace.x.push(d.log2fc);
           lvl_trace.y.push(y_log);
           lvl_trace.text.push(`<b>${d.target}</b><br>Level: ${d.level}<br>Log2 Resp: ${y_log.toFixed(2)}<br>FC: ${d.log2fc.toFixed(2)}`);
           lvl_trace.ids.push(`${d.target}_${d.level}`);
           
           lvl_trace.error_y.array.push(ucl_log - y_log);
           lvl_trace.error_y.arrayminus.push(y_log - lcl_log);
        });
        
        traces.push(lvl_trace);
    });

    if(highlight_list.length > 0) {
       const highlightData = means_search.filter(d => highlight_list.includes(d.target) && d.log2fc != null);
       
        traces.push({
        x: highlightData.map(d => d.log2fc),
        y: highlightData.map(d => Math.log2(d.response)),
        mode: 'markers+text',
        type: 'scatter', 
        name: 'Highlights',
        text: highlightData.map(d => d.target),
        textposition: 'top center',
        textfont: { color: 'black', size: 11, weight: 'bold' },
        marker: { color: 'black', size: 10, symbol: 'circle-open', line: {width: 2} },
        hoverinfo: 'skip',
        cliponaxis: false
      });
    }

    const layout = {
      height: 500,
      margin: { t: 30, b: 50, l: 60, r: 20 },
      hovermode: 'closest',
      title: { text: 'Marginal Means vs Fold Change', font: {size: 14} },
      xaxis: { 
        title: 'Log2 Fold Change (Contrast)', 
        zeroline: true, 
        zerolinecolor: '#eee',
        gridcolor: '#f9f9f9'
      },
      yaxis: { 
        title: 'Log2 Expression (Response)', 
        gridcolor: '#f9f9f9',
        zeroline: false
      },
      legend: { orientation: 'h', y: -0.15 },
      showlegend: true
    };

    const div = document.createElement("div");
    Plotly.newPlot(div, traces, layout, { responsive: true, displayModeBar: true });
    yield div;
  }
}
```


::::

### Column {width=50%}

```{ojs}
//| echo: false
//| content: card-toolbar
//| input: current_means_data
viewof means_search = Inputs.search(current_means_data, {placeholder: "Additional Target Filtering"})

```

:::: {.card title="Details"}

```{ojs}
//| echo: false
//| input:  current_means_data

viewof means_table = Inputs.table(means_search, {
  multiple: true,
  value: combined_selection, // Pre-check the combined selection
  rows: 20,
  format: {
    response: sparkbar(d3.max(means_search, d => d.response))
  }})

```


```{ojs}
//| echo: false
//| output: false
means_data_js = {
  const payload = study_data_payload;
  const lvl = means_level; 
  
  if (!payload || !lvl || lvl.length === 0 || payload.mode === "local") return null;

  if (payload.mode === "remote_js_bridge") {
    try {
      const conn = await duck_db_client;

      const formattedLevels = lvl.map(d => `'${d}'`).join(',');

      const query = `
        SELECT level, target, response, "asymp.LCL" as LCL, "asymp.UCL" as UCL
        FROM '${payload.url_emmeans}' 
        WHERE level IN (${formattedLevels})
      `;

      const result = await conn.query(query);
      return result.toArray().map(row => row.toJSON());
    } catch (e) {
      console.error("Mean Fetch Error:", e);
      return [];
    }
  }
}
```

```{webr}
#| echo: false
#| autorun: true
#| input: 
#|    - study_data_payload
#|    - means_data_js
#|    - means_level
#|    - means_highlight
#|    - means_show_labeled
#|    - selected_genes
#| define: 
#|    - current_means_data

if (!is.null(study_data_payload) && !is.null(means_level)) {
  current_means_data <- NULL
  
  if (study_data_payload$mode == "remote_js_bridge") {
     if(!is.null(means_data_js)){
       current_means_data <- jsonlite::fromJSON(jsonlite::toJSON(means_data_js))
       current_means_data <- as.data.frame(current_means_data)
     }
  } else {
    if (dbExistsTable(con, "emmeans_data")) {

        lvl_in_clause <- paste0("'", means_level, "'", collapse = ", ")
        
        query <- paste0(
          "SELECT level, target, response, \"asymp.LCL\" as LCL, \"asymp.UCL\" as UCL ",
          "FROM emmeans_data WHERE level IN (", lvl_in_clause, ")"
        )
        
        current_means_data <- dbGetQuery(con, query)
      }
  }
  
  if(!is.null(current_means_data)){
    # 1. Join with Volcano Data to get log2fc AND p.value (FIXED: Use p.value)
    if(exists("current_volcano_data") && !is.null(current_volcano_data) && nrow(current_volcano_data) > 0){

       volcano_cols <- current_volcano_data %>% 
          select(target, log2fc, p.value) %>% 
          distinct(target, .keep_all = TRUE)
          
       current_means_data <- current_means_data %>% 
          left_join(volcano_cols, by="target")
    } else {
       current_means_data$log2fc <- NA_real_
       current_means_data$p.value <- NA_real_
    }

    target_order <- current_means_data %>%
      group_by(target) %>%
      summarise(avg_resp = mean(response, na.rm = TRUE)) %>%
      arrange(avg_resp) %>%
      mutate(idx = row_number()) %>%
      select(target, idx)

    current_means_data <- current_means_data %>%
      left_join(target_order, by = "target")
  
  }
}
```

::::

# Spatial Plots

## Row {height=100%}

### Column {width=25%}

::: {.card title="Spatial Controls"}



```{ojs}
//| echo: false
viewof spatial_options = Inputs.form({
  pt_size:    Inputs.range([0.5, 10], {value: 2, step: 0.5, label: "Point Size"}),
  pt_opacity: Inputs.range([0.1, 1], {value: 0.7, step: 0.1, label: "Opacity"}),
  bg_color:   Inputs.color({label: "Background Color", value: "#e0e0e0"}),
  
  // Scale Bar
  scale_len:   Inputs.number({label: "Scale Bar (mm)", value: 1, step: 0.1}),
  scale_color: Inputs.color({label: "Scale Color", value: "#000000"}),
  show_scale:  Inputs.toggle({label: "Show Scale Bar", value: true})
})

viewof log_scale_opt = {
  if(!spatial_data || spatial_data.length === 0) return md``;
  
  const val = spatial_data[0].group_id;
  
  if (typeof val === 'number') {
    return Inputs.toggle({label: "Log Scale Color (n+1)", value: false});
  } else {
    return Inputs.toggle({label: "Log Scale Color (n+1)", value: false, disabled: true});
  }
}

spatial_colors = {
   return {
     l1: means_color_1 || "#e15759",
     l2: means_color_2 || "#4e79a7",
     other: spatial_options.bg_color
   }
}
```

<br>

**Metadata Filtering**
```{ojs}
//| echo: false
// 1. Column Selection
viewof filter_col = Inputs.select(spatial_meta_columns, {label: "Filter By:", value: null, multiple: false})

is_numeric_filter = {
  if(!spatial_data || !filter_col) return false;
  const val = spatial_data[0][filter_col];
  return typeof val === 'number';
}

viewof filter_vals = {
  if(!filter_col || !spatial_data) return md`_Select a column..._`;
  
  if (is_numeric_filter) {

    const extent = d3.extent(spatial_data, d => d[filter_col]);
    
    return Inputs.form({
      min: Inputs.number({label: "Min", value: extent[0]}),
      max: Inputs.number({label: "Max", value: extent[1]})
    });
    
  } else {

    const unique_vals = [...new Set(spatial_data.map(d => d[filter_col]))].sort();
    
    return Inputs.select(unique_vals, {
      label: "Include Values:", 
      multiple: true,
      sort: true,
      unique: true
    });
  }
}
```

<br>
**Stats**
```{ojs}
//| echo: false
md`
- **Total Loaded**: ${spatial_data ? spatial_data.length.toLocaleString() : "..."}
- **Visible**: ${spatial_filtered ? spatial_filtered.length.toLocaleString() : "..."}
`
```

:::

### Column {width=75%}



```{ojs}
//| echo: false
//| content: card-toolbar

viewof slide_selection = {
  if (!available_slides || available_slides.length === 0) return md`_Loading slides..._`;

  return Inputs.select(available_slides, {
    label: "Available Slides:",
    multiple: false,
    size: 3
  });
}

viewof term_copy2 = {
  
  if (!available_terms || available_terms.length === 0) return md`_Loading..._`;

  const mirror_input3 = Inputs.select(available_terms, {
    label: "Available Terms:",
    multiple: false,
    size: 3
  });

  return Inputs.bind(mirror_input3, viewof term_selection);
}

viewof contrast_copy2 = {
  if (!available_contrasts || available_contrasts.length === 0) return md`_Loading..._`;

  const mirror_input2 = Inputs.select(available_contrasts, {
    label: "Available Contrasts:",
    format: d => d.replace(" / ", " vs. "),
    multiple: false,
    size: 3
  });

  return Inputs.bind(mirror_input2, viewof contrast_selection);
}
```


```{ojs}
//| echo: false
//| output: false

spatial_data = {
  const payload = study_data_payload;
  const r_push = r_spatial_data_export; // Reactive link to the webR block above
  
  if (!payload) return null;

  // 1. LOCAL MODE: Always use the R-side DuckDB data
  if (payload.type === "local") {
    if (r_push && r_push.length > 0) {
       console.log("📍 Spatial Data: Received via R-Side DuckDB Bridge");
       return r_push;
    }
    // Return empty array to clear the canvas during transitions
    return []; 
  }

  // 2. REMOTE MODE: Keep the URL-based JS-bridge for manifest studies
  const term = term_selection;
  const slide = slide_selection;
  if (!term || slide === null) return null;

  try {
    const conn = await duck_db_client;
    const result = await conn.query(`SELECT * FROM "${payload.url_meta}" WHERE slide_id_numeric = ${slide}`);
    const rows = result.toArray().map(row => row.toJSON());
    rows.forEach(r => { r.group_id = r[term]; });
    return rows;
  } catch (e) {
    console.error("❌ Remote Spatial Fetch Error:", e);
    return [];
  }
}

spatial_meta_columns = {
  if (!spatial_data || spatial_data.length === 0) return [];
  
  // Dynamically grab columns from whatever study is currently loaded
  const keys = Object.keys(spatial_data[0]);
  
  return keys.filter(k => 
    !["x_slide_mm", "y_slide_mm", "group_id", "__index_level_0__", "slide_id_numeric"].includes(k)
  );
}

spatial_filtered = {
  if(!spatial_data) return [];

  if(!filter_col || !filter_vals) return spatial_data;
  
  // LOGIC SWITCH:
  if(is_numeric_filter) {
     // Numeric Filter: filter_vals is an object {min: x, max: y}
     const { min, max } = filter_vals;
     return spatial_data.filter(d => d[filter_col] >= min && d[filter_col] <= max);
  } else {
     // Categorical Filter: filter_vals is an array of selected strings
     if(filter_vals.length === 0) return spatial_data;
     return spatial_data.filter(d => filter_vals.includes(d[filter_col]));
  }
}

```

```{ojs}
//| echo: false
//| padding: 0px

{
  const data = spatial_filtered;
  const levels = means_level; 
  const opts = spatial_options;
  const colors = spatial_colors;
  const width = 1200; 
  const height = 900; 
  
  const useLog = log_scale_opt === true; 
  
  //if (!data || data.length === 0) return md`_Loading spatial data..._`;

  const isMock = !data || data.length === 0 || 
                 (data.length === 4 && d3.max(data, d => d.x_slide_mm) <= 1);
  
  if (isMock) {
    return html`
      <div style="height: 600px; display: flex; flex-direction: column; align-items: center; justify-content: center; background: #fdfdfd; border: 2px dashed #dee2e6; border-radius: 12px; text-align: center; padding: 40px;">
        <div style="font-size: 3em; opacity: 0.4;">📍</div>
        <h3 style="color: #3C3F40; margin-top: 15px;">Spatial Plot Unavailable</h3>
        <p style="max-width: 440px; color: #888; font-size: 0.9em; line-height: 1.6;">
          This study (AtoMx SIP or smiDE) does not contain any cell-level spatial metadata to display.
        </p>
        <div style="margin-top: 20px; font-size: 0.75em; border-top: 1px solid #eee; padding-top: 15px; font-style: italic; color: #b44697;">
          R users: Use <code>create_mock_sp_data()</code> to generate placeholder coordinates.
        </div>
      </div>
    `;
  }

  const isContinuous = typeof data[0].group_id === 'number';
  
  let colorScale = null;
  let termExtent = [0, 1];
  
  if(isContinuous) {
     termExtent = d3.extent(data, d => d.group_id);
     
     if (useLog) {
        // Log Scale: Map Log(Min+1) -> Log(Max+1)
        const logMin = Math.log(termExtent[0] + 1);
        const logMax = Math.log(termExtent[1] + 1);
        
        colorScale = d3.scaleSequential(d3.interpolateRgb(colors.l2, colors.l1))
          .domain([logMin, logMax]);
          
     } else {
        // Linear Scale
        colorScale = d3.scaleLinear()
          .domain(termExtent)
          .range([colors.l2, colors.l1]);
     }
  }

  const xExtent = d3.extent(data, d => d.x_slide_mm);
  const yExtent = d3.extent(data, d => d.y_slide_mm);
  
  const xMin = xExtent[0], xMax = xExtent[1];
  const yMin = yExtent[0], yMax = yExtent[1];
  
  const dataWidth = xMax - xMin;
  const dataHeight = yMax - yMin;
  
  const xScaleFactor = width / dataWidth;
  const yScaleFactor = height / dataHeight;
  const scale = Math.min(xScaleFactor, yScaleFactor) * 0.95; 
  
  const cx = width / 2;
  const cy = height / 2;
  const dx = (xMin + xMax) / 2;
  const dy = (yMin + yMax) / 2;

  const project = (x, y) => {
    return [
      cx + (x - dx) * scale,
      cy - (y - dy) * scale 
    ];
  }
  
  const pixelsPerMm = scale;

  // --- 3. DOM Setup ---
  const div = document.createElement("div");
  div.style.position = "relative";
  div.style.width = "100%";
  div.style.maxWidth = "100%";
  div.style.height = "auto";
  div.style.border = "1px solid #eee";
  div.style.background = "#fff";

  const canvas = document.createElement("canvas");
  canvas.width = width;
  canvas.height = height;
  canvas.style.width = "100%"; // Responsive CSS
  canvas.style.height = "auto";
  canvas.style.display = "block";
  div.appendChild(canvas);
  
  // Legend 
  const legend = document.createElement("div");
  legend.style.position = "absolute";
  legend.style.top = "10px";
  legend.style.right = "10px";
  legend.style.background = "rgba(255, 255, 255, 0.9)";
  legend.style.padding = "8px";
  legend.style.borderRadius = "4px";
  legend.style.border = "1px solid #ccc";
  legend.style.fontSize = "12px";
  legend.style.fontFamily = "sans-serif";
  
  if (isContinuous) {
     const label = useLog ? `Log(${term_selection} + 1)` : term_selection;
     const minLabel = useLog ? Math.log(termExtent[0] + 1).toFixed(2) : termExtent[0].toFixed(2);
     const maxLabel = useLog ? Math.log(termExtent[1] + 1).toFixed(2) : termExtent[1].toFixed(2);
     
     legend.innerHTML = `
       <strong>${label}</strong><br>
       <div style="display:flex; align-items:center; gap:5px; margin-top:4px;">
         <span>${minLabel}</span>
         <div style="width:60px; height:10px; background: linear-gradient(to right, ${colors.l2}, ${colors.l1});"></div>
         <span>${maxLabel}</span>
       </div>
     `;
  } else {
     const L1 = (levels && levels.length > 0) ? levels[0] : "Level 1";
     const L2 = (levels && levels.length > 1) ? levels[1] : "Level 2";
     legend.innerHTML = `
       <strong>${term_selection}</strong><br>
       <div style="display:flex; align-items:center; gap:5px; margin-top:2px;">
         <span style="display:inline-block; width:10px; height:10px; background:${colors.l1};"></span> ${L1}<br>
       </div>
       <div style="display:flex; align-items:center; gap:5px; margin-top:2px;">
         <span style="display:inline-block; width:10px; height:10px; background:${colors.l2};"></span> ${L2}
       </div>
     `;
  }
  div.appendChild(legend);

  const ctx = canvas.getContext("2d");
  let transform = d3.zoomIdentity;

  function draw() {
    ctx.save();
    ctx.clearRect(0, 0, width, height);
    
    ctx.translate(transform.x, transform.y);
    ctx.scale(transform.k, transform.k);
    
    const radius = opts.pt_size / transform.k;
    ctx.globalAlpha = opts.pt_opacity;

    const cOther = colors.other;
    const cL1 = colors.l1;
    const cL2 = colors.l2;
    const L1 = (levels && levels.length > 0) ? levels[0] : null;
    const L2 = (levels && levels.length > 1) ? levels[1] : null;

    for (let i = 0; i < data.length; i++) {
      const d = data[i];
      let fill = cOther;
      
      if (isContinuous) {
         if (d.group_id != null) {
            const val = useLog ? Math.log(d.group_id + 1) : d.group_id;
            fill = colorScale(val);
         }
      } else {
         if (L1 && d.group_id === L1) fill = cL1;
         else if (L2 && d.group_id === L2) fill = cL2;
      }
      
      const xy = project(d.x_slide_mm, d.y_slide_mm);
      
      ctx.beginPath();
      ctx.fillStyle = fill;
      ctx.fillRect(xy[0] - radius, xy[1] - radius, radius*2, radius*2);
    }
    
    ctx.restore();
    
    if (opts.show_scale) {
       const barLenMm = opts.scale_len;
       const barLenPx = barLenMm * pixelsPerMm * transform.k; 
       
       const margin = 20;
       const barX = width - margin - barLenPx;
       const barY = height - margin;
       
       ctx.save();
       ctx.beginPath();
       ctx.strokeStyle = opts.scale_color;
       ctx.lineWidth = 4;
       ctx.moveTo(barX, barY);
       ctx.lineTo(barX + barLenPx, barY);
       ctx.stroke();
       
       ctx.fillStyle = opts.scale_color;
       ctx.font = "bold 14px sans-serif";
       ctx.textAlign = "center";
       ctx.textBaseline = "bottom";
       ctx.fillText(`${barLenMm} mm`, barX + barLenPx/2, barY - 5);
       ctx.restore();
    }
  }

  draw();

  const zoom = d3.zoom()
    .scaleExtent([0.5, 50]) 
    .on("zoom", (event) => {
      transform = event.transform;
      requestAnimationFrame(draw);
    });

  d3.select(canvas).call(zoom);

  return div;
}
```

::: {.card .hidden}


```{webr}
#| echo: false
#| autorun: true
#| input: 
#|   - term_selection
#|   - slide_selection
#|   - db_ready
#| define:
#|   - r_spatial_data_export

r_spatial_data_export <- NULL

if (db_ready && !is.null(term_selection) && !is.null(slide_selection)) {
  # Query the view we already established during study loading
  query <- paste0(
    "SELECT *, ", term_selection, " as group_id ",
    "FROM cell_meta WHERE slide_id_numeric = ", slide_selection
  )
  
  tryCatch({
    # Fetch from the existing DuckDB connection
    r_spatial_data_export <- dbGetQuery(con, query)
  }, error = function(e) {
    cat("Spatial query waiting for view initialization...\n")
    r_spatial_data_export <- NULL
  })
}
```

:::


# Convert Your AtoMx SIP Results {scrolling="true" #sec-convert-atomx}

## Row {height=20%}

::: {.card title="How to use this dashboard with your results derived from AtoMx SIP"}

In AtoMx SIP, when your differential expression (DE) results are ready you can 
download them. This download will be a zip file named something like `output_results.zip`. 
This file is relatively large but luckily for this "Byte-sized" dashboard we only
need a few of the columns. You can use this section to format that large zip file
into a dashboard-ready zip file that contains essential information. See the card
below to load your AtoMx-derived DE zip file. You'll be given the option to add
optional study description and model formula that can be handy when you are comparing
multiple studies at once. When the processing is finished, you will have the option
to download the dashboard-ready and formatted zip file, load it directly into this
dashboard, or both. One thing to note is that the DE zip file that is downloaded
from AtoMx SIP does not contain cell-level metadata and so the Spatial Plots section
will only contain a placeholder dataset (just four 'cells') and not actual data.

:::

## Row {height=30%}

### Column {width=50%}

::: {.card title="Step 1: Load AtoMx SIP DE Results"}

```{ojs}
//| echo: false
//| output: true

viewof atomx_zip_file = Inputs.file({
  label: "Import AtoMx SIP DE Results (.zip):",
  accept: ".zip"
})

atomx_surgical_ingestor = {
  if (!atomx_zip_file) return null;
  
  try {
    console.log("🛠️ INGESTOR START: Reading ArrayBuffer...");
    const buffer = await atomx_zip_file.arrayBuffer();
    const z = await import("https://cdn.jsdelivr.net/npm/@zip.js/zip.js@2.7.34/+esm");
    
    // Create the reader from the Uint8Array of the full zip
    const reader = new z.ZipReader(new z.Uint8ArrayReader(new Uint8Array(buffer)));
    const entries = await reader.getEntries();
    
    const targets = ["emmeans.csv", "pairwise.csv", "one.vs.rest.csv", "one.vs.all.csv"];
    const conn = await duck_db_client;
    const db = conn.instance || conn.db || conn;

    for (const entry of entries) {
      const match = targets.find(t => entry.filename.toLowerCase().includes(t));
      if (match) {
        console.log(`📦 ZIP ENTRY FOUND: ${entry.filename}`);
        
        // FIX: Ensure we are awaiting the full data extraction
        const uint8 = await entry.getData(new z.Uint8ArrayWriter());
        
        if (uint8 && uint8.length > 0) {
          await db.registerFileBuffer(match, uint8);
          console.log(`✅ REGISTERED: ${match} (${uint8.length.toLocaleString()} bytes)`);
        } else {
          console.warn(`⚠️ WARNING: ${match} extracted as 0 bytes!`);
        }
      }
    }
    
    await reader.close();
    return { status: "ready", timestamp: Date.now() };
  } catch (err) {
    console.error("❌ INGESTOR CRITICAL ERROR:", err);
    return { status: "error", message: err.message };
  }
}
```


```{ojs}
//| echo: false
//| output: false

atomx_parquet_generator = {
  if (!atomx_surgical_ingestor || atomx_surgical_ingestor.status !== "ready") return null;
  if (!atomx_metadata || atomx_metadata.process_btn === 0) return null;
  
  const conn = await duck_db_client;
  console.log("🪄 GENERATOR START: Building SQL Views...");

  try {
    // FIX: Use DuckDB's internal schema field mapping
    console.log("🔍 Probing columns in pairwise.csv...");
    const schemaRes = await conn.query("SELECT * FROM read_csv_auto('pairwise.csv') LIMIT 0");
    const cols = schemaRes.schema.fields.map(f => f.name);
    console.log("📋 Detected Columns:", cols);
    
    const n1 = cols.includes("ncells_1") ? "ncells_1" : "NULL as ncells_1";
    const n2 = cols.includes("ncells_2") ? "ncells_2" : "NULL as ncells_2";

    // 1. Harmonize EMMEANS
    await conn.query(`
      CREATE OR REPLACE VIEW emmeans_final AS
      WITH base AS (
        SELECT *,
          regexp_replace(level, term, '') as suffix_str,
          TRY_CAST(regexp_replace(level, term, '') AS DOUBLE) as suffix_val,
          (level LIKE term || '%') AND (TRY_CAST(regexp_replace(level, term, '') AS DOUBLE) IS NOT NULL) as is_continuous
        FROM read_csv_auto('emmeans.csv')
        WHERE term != 'otherct_expr'
      )
      SELECT 
        term,
        CASE 
          WHEN is_continuous AND category = 'c1' THEN 'mean + 1SD (' || printf('%.2f', suffix_val) || ')'
          WHEN is_continuous AND category = 'c2' THEN 'mean  (' || printf('%.2f', suffix_val) || ')'
          ELSE level 
        END as level,
        target, response, "asymp.LCL", "asymp.UCL"
      FROM base
    `);

    // 2. Build Master PAIRWISE Union
    await conn.query(`
      CREATE OR REPLACE VIEW pairwise_final AS
      SELECT 'pairwise' as contrast_type, term, contrast, target, log2(fold_change) as log2fc, "p.value", ${n1}, ${n2} FROM read_csv_auto('pairwise.csv')
      UNION ALL
      SELECT 'one_vs_rest', term, contrast, target, log2(fold_change), "p.value", ${n1}, ${n2} FROM read_csv_auto('one.vs.rest.csv')
      UNION ALL
      SELECT 'one_vs_all', term, contrast, target, log2(fold_change), "p.value", ${n1}, ${n2} FROM read_csv_auto('one.vs.all.csv')
    `);

    console.log("✅ GENERATOR SUCCESS!");
    return { ready: true, timestamp: Date.now() };

  } catch (err) {
    console.error("❌ GENERATOR SQL ERROR:", err);
    return { ready: false, error: err.message };
  }
}
```

:::

### Column {width=50%}

::: {.card title="Step 2: Add Additional Information"}


```{ojs}
//| echo: false
//| output: true
viewof atomx_metadata = {
  
  if (!atomx_surgical_ingestor || atomx_surgical_ingestor.status !== "ready") {
    return html`
      <div style="margin-top: 10px; padding: 15px; border: 1px dashed #ddd; border-radius: 8px; background: #fafafa; color: #666;">
        <h5 style="text-transform: uppercase; color: #888; font-size: 0.75em; margin-bottom: 8px;">AtoMx Ingestor Status</h5>
        <div style="font-size: 0.9em; display: flex; align-items: center; gap: 10px;">
          <div class="spinner-border spinner-border-sm text-secondary" role="status"></div>
          <span>Waiting for AtoMx SIP results import from Step 1...</span>
        </div>
      </div>
    `;
  }

  
  const form = Inputs.form({
    study_name:  Inputs.text({label: "Study Name", value: "AtoMx Analysis"}),
    description: Inputs.text({label: "Description", value: "AtoMx SIP Export"}),
    formula:     Inputs.text({label: "Model Formula", placeholder: "e.g. ~annotated_domain"}),
    process_btn: Inputs.button("🚀 Format Results", {value: 0})
  });


  form.style.border = "1px solid #eee";
  form.style.borderRadius = "8px";
  form.style.padding = "20px";
  form.style.background = "#fff";
  form.style.marginTop = "10px";

  // Prepend the header inside the form card
  const header = html`
    <h5 style="text-transform: uppercase; color: #888; font-size: 0.75em; margin-bottom: 12px; border-bottom: 1px solid #eee; padding-bottom: 8px;">
      Optional Study Metadata
    </h5>
  `;
  form.prepend(header);

  return form;
}
```

:::

## Row {height=50%}

::: {.card title="Step 3: Choose What To Do With Formatted Results"}


<div id="atomx-download-container" style="margin-top: 20px; padding: 15px; border: 1px solid #ddd; border-radius: 8px; background: #fafafa;">
  </div>

```{ojs}
//| echo: false
//| output: false

atomx_downloader_manager = {
  const container = d3.select("#atomx-download-container");
  
  // Guard: Hide container if data isn't processed yet
  if (!atomx_parquet_generator || !atomx_parquet_generator.ready) {
    container.style("display", "none");
    return null;
  }

  container.style("display", "block").html(""); // Clear and show


  const ui = html`
    <h5 style="margin-top: 0;">🚀 Formatting Complete</h5>
    <p style="font-size: 0.9em; color: #555;">Choose how you would like to proceed with your harmonized study:</p>
    
    <div class="d-grid gap-2 d-md-block">
      <button id="btn-atomx-download-only" class="btn btn-outline-primary">💾 Download Only</button>
      <button id="btn-atomx-launch-only" class="btn btn-outline-success">⚡ Launch Only</button>
      <button id="btn-atomx-both" class="btn btn-primary">🚀 Download & Launch</button>
    </div>
    
    <div id="atomx-download-status" style="margin-top: 12px; font-weight: 500;"></div>
  `;


  container.node().appendChild(ui);


  const status = ui.querySelector("#atomx-download-status");
  const btnDownload = ui.querySelector("#btn-atomx-download-only");
  const btnLaunch = ui.querySelector("#btn-atomx-launch-only");
  const btnBoth = ui.querySelector("#btn-atomx-both");


  const runConversion = async () => {
      const conn = await duck_db_client;
      const db = conn.instance || conn.db || conn;
      

      await conn.query(`
        CREATE OR REPLACE VIEW pairwise_scrubbed AS 
        SELECT * FROM pairwise_final WHERE term != 'otherct_expr'
      `);
      
      await conn.query(`
        CREATE OR REPLACE VIEW emmeans_scrubbed AS 
        SELECT * FROM emmeans_final WHERE term != 'otherct_expr'
      `);
  

      const termRes = await conn.query("SELECT DISTINCT term FROM pairwise_scrubbed LIMIT 1");
      const activeTerm = termRes.toArray()[0]?.term || "Term";
      
      const levelRes = await conn.query("SELECT DISTINCT level FROM emmeans_scrubbed LIMIT 2");
      const levels = levelRes.toArray().map(row => row.level);
      const l1 = levels[0] || 'Level_A';
      const l2 = levels[1] || l1;
  
      console.log(`🧼 FILTERING: Removed 'otherct_expr'. Active Term: ${activeTerm}`);
  

      await conn.query("COPY (SELECT * FROM pairwise_scrubbed) TO 'pairwise.parquet' (FORMAT 'PARQUET')");
      await conn.query("COPY (SELECT * FROM emmeans_scrubbed) TO 'emmeans.parquet' (FORMAT 'PARQUET')");
  

      await conn.query(`
        CREATE OR REPLACE VIEW spatial_smart_placeholder AS
        SELECT 0.0 as x_slide_mm, 0.0 as y_slide_mm, 1 as slide_id_numeric, '${l1}' as "${activeTerm}" UNION ALL
        SELECT 0.0, 1.0, 1, '${l1}' UNION ALL
        SELECT 1.0, 0.0, 1, '${l2}' UNION ALL
        SELECT 1.0, 1.0, 1, '${l2}'
      `);
      await conn.query("COPY (SELECT * FROM spatial_smart_placeholder) TO 'cell_metadata.parquet' (FORMAT 'PARQUET')");
  

      const countRes = await conn.query("SELECT COUNT(DISTINCT target) as n FROM pairwise_scrubbed");
      const targetCount = Number(countRes.toArray()[0].n);
      await conn.query(`
        COPY (SELECT '${atomx_metadata.study_name}' as Name, '${atomx_metadata.description}' as Description, 
              '${atomx_metadata.formula}' as Formala, ${targetCount} as "Targets tested") 
        TO 'study_header.parquet' (FORMAT 'PARQUET')`);
  

      return {
        p: await db.copyFileToBuffer('pairwise.parquet'),
        e: await db.copyFileToBuffer('emmeans.parquet'),
        m: await db.copyFileToBuffer('cell_metadata.parquet'),
        h: await db.copyFileToBuffer('study_header.parquet')
      };
    };

  const createZipBlob = async (buffers) => {
    const z = await import("https://cdn.jsdelivr.net/npm/@zip.js/zip.js@2.7.34/+esm");
    const zipWriter = new z.ZipWriter(new z.BlobWriter("application/zip"));
    await zipWriter.add("pairwise.parquet", new z.Uint8ArrayReader(new Uint8Array(buffers.p)));
    await zipWriter.add("emmeans.parquet", new z.Uint8ArrayReader(new Uint8Array(buffers.e)));
    await zipWriter.add("cell_metadata.parquet", new z.Uint8ArrayReader(new Uint8Array(buffers.m)));
    await zipWriter.add("study_header.parquet", new z.Uint8ArrayReader(new Uint8Array(buffers.h)));
    return await zipWriter.close();
  };

  const triggerDownload = (blob) => {
    const url = URL.createObjectURL(blob);
    const a = document.createElement("a");
    a.href = url; a.download = `${atomx_metadata.study_name}_smiDE.zip`;
    document.body.appendChild(a);
    a.click();
    document.body.removeChild(a);
  };


  btnDownload.onclick = async () => {
    status.innerHTML = "⚙️ Generating Zip...";
    const buffers = await runConversion();
    const blob = await createZipBlob(buffers);
    triggerDownload(blob);
    status.innerHTML = "✅ Download Complete.";
  };

  btnLaunch.onclick = async () => {
    status.innerHTML = "🚀 Launching Study...";
    const buffers = await runConversion();
    const blob = await createZipBlob(buffers);

    mutable new_zip_data = {
      id: Date.now(),
      filename: `${atomx_metadata.study_name}_smiDE.zip`,
      data: new Uint8Array(await blob.arrayBuffer())
    };
    status.innerHTML = "✅ Study Launched! Check the sidebar.";
  };

  btnBoth.onclick = async () => {
    status.innerHTML = "⚙️ Processing Both...";
    const buffers = await runConversion();
    const blob = await createZipBlob(buffers);
    triggerDownload(blob);
    mutable new_zip_data = {
      id: Date.now(),
      filename: `${atomx_metadata.study_name}_smiDE.zip`,
      data: new Uint8Array(await blob.arrayBuffer())
    };
    status.innerHTML = "✅ Success! Zip downloaded and Study launched.";
  };

  return ui;
}
```

:::

# Convert Your smiDE Results {scrolling="true" #sec-convert-smide}

::: {.card title="How to use this dashboard with your results derived from smiDE"}

You can use this dashboard to explore your own `smiDE` results. For now this
dashboard only works using differential expression results obtained directly 
from the `smiDE` R package. So far I have tested it with two cases: a discrete/
categorical grouping variable and a continuous one. 

Let's say you have a differential expression object obtained from calling the 
`smiDE::smi_de` function:

```{r}
#| echo: true
#| eval: false

library(smiDE)

# Example function call below that examines tumor cell expression across
# different spatial domains (annotated_domains). 
de_obj <- 
   smi_de(assay_matrix = assay_matrix_use
          ,metadata = meta_use
          ,formula = ~RankNorm(otherct_expr) + annotated_domain + offset(log(nCount_RNA)) 
          ,pre_de_obj = pre_de_obj
          ,neighbor_expr_cell_type_metadata_colname = "celltype_broad"
          ,neighbor_expr_overlap_weight_colname = NULL
          ,neighbor_expr_overlap_agg ="sum"
          ,neighbor_expr_totalcount_normalize = TRUE
          ,neighbor_expr_totalcount_scalefactor =  tc_scalefactors 
          ,family="nbinom2"
          ,cellid_colname = "cell_id_numeric"
          ,targets=genes_to_analyze,
          nCores=30
   ) 
res_list <- results(de_obj)
```

Since the output structure differs slighlty depending on whether your DE results
are based on discrete or continous terms, we'll use this harmonizing function
to format the results into a single format that the dashboard can understand. 
Specifically, the function below converts the _pairwise results_ and the 
_emmeans results_. 
```{webr}
#| echo: true
#| eval: false

library(data.table)
library(dplyr)
library(stringr)

#' Harmonize smiDE Results for Dashboard
#'
#' @param res_list The output list directly from smiDE::results(de_obj)
#' @return A list containing 'pairwise' and 'emmeans' data.frames with synchronized continuous labels
harmonize_de_results <- function(res_list) {
  
  emmeans_raw <- res_list$emmeans
  if ("term" %in% names(emmeans_raw)) {
    emmeans_raw <- emmeans_raw[term != "otherct_expr"]
  }
  
  emmeans_processed <- emmeans_raw %>%
    mutate(
      # Extract numeric suffix from level (e.g., "term42.697" -> "42.697")
      suffix_str = str_remove(level, fixed(term)),
      suffix_val = suppressWarnings(as.numeric(suffix_str)),
      
      is_continuous = !is.na(suffix_val) & str_starts(level, fixed(term))
    ) %>%
    mutate(
      clean_level = case_when(
        is_continuous ~ {
          direction <- case_when(
            category == "c1" ~ "+ 1SD",
            category == "c2" ~ "",
            TRUE ~ ""
          )
          paste0("mean ", direction, " (", sprintf("%.2f", suffix_val), ")")
        },
        TRUE ~ level
      ),
      
      lookup_key = ifelse(is_continuous, 
                          paste0(term, "_", sprintf("%.2f", suffix_val)), 
                          NA_character_)
    )
  
  continuous_map <- emmeans_processed %>%
    filter(is_continuous) %>%
    select(lookup_key, clean_level) %>%
    distinct() %>%
    tibble::deframe() 
  
  emmeans_out <- emmeans_processed %>%
    select(term, level = clean_level, target, response, asymp.LCL, asymp.UCL) %>%
    as.data.frame()
  
  pairwise_combined <- rbindlist(
    res_list[c("pairwise", "one.vs.rest", "one.vs.all")], 
    idcol = "contrast_type", 
    use.names = TRUE, 
    fill = TRUE 
  )
  
  if ("term" %in% names(pairwise_combined)) {
    pairwise_combined <- pairwise_combined[term != "otherct_expr"]
  }
  
  if (!"ncells_1" %in% names(pairwise_combined)) pairwise_combined[, ncells_1 := NA]
  if (!"ncells_2" %in% names(pairwise_combined)) pairwise_combined[, ncells_2 := NA]
  
  pairwise_out <- pairwise_combined %>%
    mutate(
      log2fc = log2(fold_change),
      
      is_continuous_row = is.na(ncells_1) & is.na(ncells_2)
    ) %>%
    rowwise() %>%
    mutate(
      contrast = if (is_continuous_row) {
        parts <- str_split(contrast, " / ", simplify = TRUE)

        val1 <- as.numeric(str_remove(parts[1], fixed(term)))
        val2 <- as.numeric(str_remove(parts[2], fixed(term)))
        
        key1 <- paste0(term, "_", sprintf("%.2f", val1))
        key2 <- paste0(term, "_", sprintf("%.2f", val2))
        
        label1 <- ifelse(key1 %in% names(continuous_map), continuous_map[key1], parts[1])
        label2 <- ifelse(key2 %in% names(continuous_map), continuous_map[key2], parts[2])
        
        paste0(label1, " / ", label2)
        
      } else {
        contrast
      }
    ) %>%
    ungroup() %>%
    select(term, contrast, target, log2fc, p.value, ncells_1, ncells_2) %>%
    as.data.frame()
  
  return(list(pairwise = pairwise_out, emmeans = emmeans_out))
}
```

Now convert the `res_list` into the formatted parts. 

```{r}
#| echo: true
#| eval: false

res_formatted <- harmonize_de_results(res_list)
pairwise <- res_formatted$pairwise
emmeans <- res_formatted$emmeans

```


The `pairwise` table and the `emmeans` table two of the four required components.
To create the `study_header` -- which is simply a description data set that helps
provide a little more context to the data displayed on the dashboard itself, we'll
create it like this: 

```{r}
#| echo: true
#| eval: false

study_header <- data.frame( # <1>
  `Name` = "Tumor cells across spatial domains",
  `Description` = "Colon Cancer WTX sample",
  `Formala` = "~RankNorm(otherct_expr) + annotated_domain + offset(log(nCount_RNA))",
  `Targets tested` = length(genes_to_analyze)
)
```
Note: The column names must **match exactly** as given.


And finally, it can be useful view the cells in space to get a sense of the 
spatial structure of groupings. We'll do this by selecting columns within
our cell-level metadata object. Keep in mind that this dataset is loaded fully
in your browser's memory and so very large dataframes might slow down or even
crash your browser.

In the code snippet below, I'm interested in looking at the broad cell types 
and the annotated domain. Note that there are a minimum of four required columns that
are expected in the dashboard. The first three are `x_slide_mm`,
`y_slide_mm`, and `slide_id_numeric`. These are used to facet the spatial plot.
The other required columns are the metadata columns used for your grouping variable. 
In the example above, that grouping column was named `annotated_domain` but this
can vary based on the study. These columns are used to _color_ the spatial plots
when a given contrast is selected. Other columns can optionally be added. For example,
`celltype_broad` is useful to include here is we used that column to filter our
data and in the spatial plots section we can filter the cells to include only
cells that have a `celltype_broad` value of `tumor` to match our DE analysis.

```{r}
#| echo: true
#| eval: false

# meta_df == the full cell-level metadata (i.e., obs)
meta_display <- data.table(meta_df)
meta_display <- meta_display[, .(
  x_slide_mm = sdimx, 
  y_slide_mm = sdimy, 
  celltype_broad,
  annotated_domain
)]
if('slide_id_numeric' %in% colnames(meta_df)){
  meta_display$slide_id_numeric <- meta_df$slide_id_numeric
} else {
  meta_display$slide_id_numeric <- 1L
}

```

Alternatively, if you just want the volcano plots and emmeans and want to skip 
the spatial plotting together, just create and pass a placeholder data.frame like
so:

```{r}
#| echo: true
#| eval: false

create_mock_sp_data <- function(emmeans_formatted) {
  terms <- emmeans_formatted %>% 
    select(term) %>% 
    pull()
  
  mock_template <- data.frame(
    'x_slide_mm' = c(0, 0, 1, 1),
    'y_slide_mm' = c(0, 1, 0, 1), 
    'slide_id_numeric' = rep(1L, 4)
  )
  
  for(i in seq_len(length(terms))){
    mock_template[[terms[i]]] <- rep(NA, 4)
  }
  
return(mock_template)
}
# this will just add 4 rows data which the dashboard uses as a signature for
# "Not Applicable"
meta_display <- create_mock_sp_data(emmeans)

```

At this point we are ready to package these four components so they can be loaded
in your browser's tab (specifically, in the virtual file system [VFS] that webR uses). 
The two steps to this procedure is:

1. convert data to parquet.
2. zip up all parquet files into a single zip file.

When converting data to parquet, we'll use this `write_opt_parquet` function. It's just a wrapper function
for writing to parquet but, to save memory, one could reduce the numeric precision
of columns from R's float64 to either 32 bit or 16 bit, if desired. Keep in mind
that things like p-values will likely need greater precision.

```{webr}
#| echo: true
#| eval: false
#| autorun: true
#| edit: false

library(arrow)
write_opt_parquet <- function(df, filename, f32_cols = NULL, f16_cols = NULL) {
  
  # Convert to Arrow Table
  arrow_table <- as_arrow_table(df)
  
  # Cast Float32 columns
  if (!is.null(f32_cols)) {
    # Check which requested columns actually exist in this dataframe
    valid_cols <- intersect(names(df), f32_cols)
    if(length(valid_cols) > 0) {
      arrow_table <- arrow_table %>% 
        mutate(across(all_of(valid_cols), ~ cast(.x, float32())))
    }
  }

  # Cast Float16 columns
  if (!is.null(f16_cols)) {
    valid_cols <- intersect(names(df), f16_cols)
    if(length(valid_cols) > 0) {
      arrow_table <- arrow_table %>% 
        mutate(across(all_of(valid_cols), ~ cast(.x, float16())))
    }
  }

  # Write to disk
  write_parquet(arrow_table, filename)
}


```

Write the parquet files to disk. Note that the folder names can vary but
the file names must match exactly.

```{r}
#| echo: true
#| eval: false

parquet_dir <- "./your_parquet_folder"
dir.create(parquet_dir)

write_opt_parquet(study_header, file.path(parquet_dir, "study_header.parquet"))

write_opt_parquet(
  pairwise, 
  file.path(parquet_dir, "pairwise.parquet")
)

write_opt_parquet(
  emmeans, 
  file.path(parquet_dir, "emmeans.parquet")
)

write_opt_parquet(
  meta_display,
  file.path(parquet_dir, "cell_metadata.parquet"),
  f32_cols = c('x_slide_mm', 'y_slide_mm'),
  f16_cols = NULL
)
```

And finally, zip it up! You can name the zip file whatever you like.
```{r}
#| echo: true
#| eval: false

files_to_zip <- c("study_header.parquet", 
                  "pairwise.parquet", 
                  "emmeans.parquet", 
                  "cell_metadata.parquet")

zip::zip(
  zipfile = "you_can_name_this_whatever_you_like.zip",
  files = files_to_zip, 
  root = parquet_dir 
)
```

And that should be it. You should be able to "upload" that zip file on the left 
side panel of the dashboard. 

:::

# Advanced

## Row {height="80%"}

### Column {width="100%"}

::: {.card title="Description"}
This is the code that generates the volcano plot. For advanced users who wish
to modify the volcano plot aesthetics 
beyond what is available within `ADJUST IMAGE AESTHETICS`, you can adjust the code
and click `Run Code`. This will, in turn, use your custom code reactively. If any
errors occur, you can simply click `START OVER` or simply refresh your browser, 
to return the plotting function back to its original state. You can install and use
additional R packages if you like if they are already [compiled to WebAssembly](https://repo.r-wasm.org/sembly){target="_blank"}.


. This is "living code"
is used to generate the and adjustmentused to reactively generate the plots of the dashboard. You can modify them to generate completely custom aesthetics. If run into issues, you can always press 'START OVER' to return the plotting function back to its original form.

```{webr}
#| title: Volcano Plot Source Code
#| edit: true
#| output: true
#| autorun: true
#| warning: false
#| input: 
#|   - volcano_fc_range
#|   - volcano_pval_threshold
#|   - volcano_pt_size
#|   - volcano_alpha
#|   - volcano_pt_color
#|   - volcano_sig_color
#|   - volcano_high_fc_label
#|   - volcano_low_fc_label
#|   - volcano_pval_label
#|   - volcano_include_selected_genes
#|   - selected_genes
#| define: 
#|   - volcano_plot_emit
#|   - plotVolcano
#|   - argsVolcano

volcano_plot_emit <- TRUE
if(exists(".Random.seed")){
  rm(.Random.seed)
}
set.seed(42L)
plotVolcano <- function(df, vfc, vpl, vpt, valpha, vptcolor){

  contrast_string <- gsub(" vs. ", " / ", df$contrast[1])
  levels <- unlist(strsplit(contrast_string, " / "))

  if(length(levels) == 2) {
    numerator <- levels[1]
    if("ncells_1" %in% colnames(df) && !is.na(df$ncells_1[1])){
      numerator <- paste0(numerator, " (n=",df$ncells_1[1], ")" )
    }    
    denominator <- levels[2]
    if("ncells_2" %in% colnames(df) && !is.na(df$ncells_2[1])){
      denominator <- paste0(denominator, " (n=",df$ncells_2[1], ")" )
    }  
    dynamic_label <- paste(denominator, "\u2190 log2(FC) \u2192", numerator)
  } else {
    dynamic_label <- "log2(FC)"
  }
  
  df$color_group <- ifelse(
    df$p.value < vpl & abs(df$log2fc) > vfc, 
    "Significant", 
    "Not Significant"
  )
  genes_to_label <- c()
  if(length(selected_genes)>0 && volcano_include_selected_genes=="Yes"){
    genes_to_label <- c(genes_to_label, 
                        selected_genes)
  }
  if(volcano_high_fc_label>0){
    genes_to_label <- c(genes_to_label,
                        df %>% arrange(-log2fc) %>% head(volcano_high_fc_label) %>% select(target))
  }
  if(volcano_low_fc_label>0){
    genes_to_label <- c(genes_to_label,
                        df %>% arrange(log2fc) %>% head(volcano_low_fc_label) %>% select(target))
  }  
  if(volcano_pval_label>0){
    genes_to_label <- c(genes_to_label,
                        df %>% arrange(p.value) %>% head(volcano_pval_label) %>% select(target))
  }  
  genes_to_label <- unique(unlist(genes_to_label))
  
  p <- ggplot(df, aes(x=log2fc, y=-log10(p.value))) + 
      geom_point(aes(color = color_group), 
                 size=vpt, alpha=valpha) +
      scale_color_manual(values = c("Significant" = volcano_sig_color, "Not Significant" = vptcolor)) + 
      geom_vline(xintercept = c(-vfc, vfc)) + 
      geom_hline(yintercept = -log10(vpl)) + 
      labs(x = dynamic_label) + 
      theme_bw() + 
      theme(legend.position = "none")
  
  if(length(genes_to_label)>0){
    p <- p + ggrepel::geom_label_repel(data=filter(df, target %in% genes_to_label),
                                  aes(x=log2fc, y=-log10(p.value), label=target))
  }
  stats = data.frame(
    `Category` = c(paste0("Significant at P-value ", vpl, " threshold: "),
                   paste0("Number genes with log2FC >= ",  vfc, ":"),
                   paste0("Number genes with log2FC <= ",  -vfc, ":"),
                   paste0("Signifant and high log2FC:"),
                   paste0("Signifant and low log2FC:")
                   ),
    `Value` = c(filter(df, p.value <= vpl) %>% nrow(),
                filter(df, log2fc >= vfc) %>% nrow(),
                filter(df, log2fc <= -vfc) %>% nrow(),
                filter(df, log2fc >= vfc, p.value <= vpl) %>% nrow(),
                filter(df, log2fc <= -vfc, p.value <= vpl) %>% nrow()
                )
  )
  return(list(p, stats, genes_to_label)) # Return labeled genes as the 3rd element
}

argsVolcano <- list(vfc=volcano_fc_range, 
                    vpl=volcano_pval_threshold,
                    vpt=volcano_pt_size,
                    valpha=volcano_alpha,
                    vptcolor=volcano_pt_color)
```

:::


# webR Console

This `webR` console is available for anything from quick calculations to creating additional plots. For a list of pre-compiled
R packages that are available take a look at [https://repo.r-wasm.org/](https://repo.r-wasm.org/){target="_blank"}.

```{webr}

ls()

```


<!-- This script simply changes out the temporary OJS runtime error messages  -->
<!-- that pop up when the cells are initializing.  -->
```{ojs}
//| echo: false
//| output: false

{
  const catchRuntimeLoadingErrors = () => {
    // Target the specific Quarto callout structure you captured
    const errors = document.querySelectorAll('.observablehq--error, .ojs-in-a-box-waiting-for-module-import');
    
    errors.forEach(el => {
      const callout = el.querySelector('.callout-important');
      if (callout) {
        callout.style.display = 'none'; 
        
        if (!el.querySelector('.custom-loading-label')) {
          const label = document.createElement('div');
          label.className = 'custom-loading-label';
          label.innerHTML = `
            <div style="padding: 2rem; border: 1px dashed #ccc; border-radius: 8px; 
                        background: #f9f9f9; color: #888; text-align: center; font-style: italic;">
              ⌛ Initializing Dashboard Components...
            </div>`;
          el.prepend(label);
        }
      }
    });
  };

  catchRuntimeLoadingErrors();

  const observer = new MutationObserver(catchRuntimeLoadingErrors);
  observer.observe(document.body, { childList: true, subtree: true });

  invalidation.then(() => observer.disconnect());
}
```