---
title: "CellPoly quick start vignette"
output: 
rmarkdown::html_vignette: 
toc: true
fig_width: 7 
fig_height: 8 
vignette: >
  %\VignetteIndexEntry{Straightforward CellPoly run}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<style>
p.caption {
  font-size: 1.5em;
}
</style>
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Installation
```{r installation, eval=FALSE}
# method 1
# devtools::install_github("Nanostring-Biostats/CosMx-Analysis-Scratch-Space/_code/cellPoly@Main")

# method 2
# setwd() in Rstudio to the CellPoly directory
# run devtools::install()

```   

### Overview

The CellPoly package is designed to facilitate plotting cells' polygonal boundaries 
while working in an R session. For each cell, it finds a polygon encompassing its transcripts. This will diverge from segmentation masks, but it will place each transcript
in the same cell (ignoring 3-d considerations). 

The only necessary input is a data frame of transcript positions, along with the cell IDs 
they are assigned to. 

This vignette demonstrates the use of the CellPoly package to plot [CosMx](https://www.nature.com/articles/s41587-022-01483-z) data. CosMx is a spatial molecular imager for sub-cellular resolution of mRNA transcripts and protein targets. We analyze a subset of a lupus nephritis experiment. 

### Package architecture & workflow:

CellPoly works on a named list of polygons, which each element corresponding to
a unique cell ID. 
To begin, the elements of this list are empty. 
Cell polygons can be calculated up-front using the cellPolys() function,
or they can be computed on the fly while drawing plots using the drawPolys() function.
The workflow can be arranged such that only the polygons needed for a given plot are computed, and so no cell's polygon need be computed twice. 
As code{cellPolys} or \code{drawPolys} (which itself calls \code{cellPolys}) is run, any polygons they compute get added to the ever-growing polygons list. 

### Preliminaries
First, we load the package and access the example data:
```{r setup}
set.seed(0)

library(CellPoly)
data("cosmx_nephritis")
str(cosmx_nephritis)
transcript_df = cosmx_nephritis$transcript_df
```

### The CellPoly workflow

The only input we need is transcript positions:

```{r tdf}
head(transcript_df)
```

Note that the CellPoly functions expect the colnames "cell_id", "x" and "y" to be present.

For this dataset, we also have pre-defined cell types and cell type colors:
```{r celldata}
frame()
legend("center",
       pch = 16, 
       col = cosmx_nephritis$cellcols, 
       legend = names(cosmx_nephritis$cellcols),
       cex = 0.4)

```

We'll start by initializing an empty polygons object, with elements for each 
cell in the transcript data frame:

```{r init}
polys <- initPolys(cell_ids = unique(transcript_df$cell_id))
polys[1:3]
```

We can calculate all polygons at this point:

```{r runcellpolys}
polys <- cellPolys(polys = polys,
                   transcript_df = transcript_df,
                   type = "chull",   # faster chull, slower/buggy "ahull" also available
                   cell_ids = NULL)

polys[1]
```

Now we can plot polygons:

```{r drawpolys1}
drawPolys(polys, 
          cell_ids = NULL, #tempids, 
          col = scales::alpha(cosmx_nephritis$cellcols[cosmx_nephritis$celltype[names(polys)]], 0.5),
          border = cosmx_nephritis$cellcols[cosmx_nephritis$celltype[names(polys)]], 
          add = FALSE, outputpolys = FALSE, xaxt = "n", yaxt = "n", xlab = "", ylab = "", asp = 1)
```

The assignment back to polys is important, since drawPolys returns the polygons list. 
(This is because it can add to the polygons list if it was incomplete to begin with.)


To finish this example, you might want to plot a couple genes across this view:

```{r plotgenes}
drawPolys(polys, 
          cell_ids = NULL, #tempids, 
          col = scales::alpha(cosmx_nephritis$cellcols[cosmx_nephritis$celltype[names(polys)]], 0.5),
          border = cosmx_nephritis$cellcols[cosmx_nephritis$celltype[names(polys)]], 
          add = FALSE, outputpolys = FALSE, xaxt = "n", yaxt = "n", xlab = "", ylab = "", asp = 1)
# which genes to show and their colors:
genecols = c("IFITM1" = "red", 
             "IFITM3" = "blue") 
# points for genes:
for (gene in names(genecols)) {
  points(transcript_df[transcript_df$target == gene, c("x", "y")], 
         col = genecols[gene], 
         pch = 16, cex = 0.2)
}
legend("bottomleft", col = genecols, legend = names(genecols), pch = 16, cex = 0.9)


```

### Inattentive workflow

You can also work with just the drawPolys() function, skipping the cellPolys call:

```{r inattentive}
#initialize:
polys <- initPolys(cell_ids = unique(transcript_df$cell_id))

# draw a few polygons, saving them so they don't need to be recalcualted
polys <- drawPolys(polys, 
                   cell_ids = names(polys)[c(1,200, 400, 700, 900)], 
                   transcript_df = transcript_df,
                   col = "#00008B80", border = "#00008BFF", 
                   add = FALSE, 
                   outputpolys = TRUE,
                   type = "chull",
                   xaxt = "n", yaxt = "n", xlab = "", ylab = "", asp = 1)

# the polys object now have values for the cell ids we specified:
polys[1]
# but the other cells still have no polys:
polys[2]

# now draw the rest:
polys <- drawPolys(polys, 
                   cell_ids = NULL,
                   transcript_df = transcript_df,
                   col = "#00008B80", border = "#00008BFF", 
                   add = FALSE, 
                   outputpolys = TRUE,
                   type = "chull",
                   xaxt = "n", yaxt = "n", xlab = "", ylab = "", asp = 1)
polys[2]

```


### Session Info

```{r sessioninfo}
sessionInfo()
```
