---
title: "InSituTree hierarchical cell typing"
author: "Claire Williams"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{InSituTree hierarchical cell typing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

The `InSituTree` package is a small package that builds on [`InSituType`](https://github.com/Nanostring-Biostats/InSituType/tree/main) and 
provides a specialized application for its use. With this approach, you can 
perform supervised cell typing in a hierarchical manner -- first making broad 
cell type assignments and then making progressively finer classifications.

## Installation

```{r installation, eval=FALSE}
# Install from github
devtools::install_github(
  "https://github.com/Nanostring-Biostats/CosMx-Analysis-Scratch-Space",
  subdir = "_code/InSituTree")
```

## Data preparation

Load the package as well as example data. For the example data, we use a small
CosMx SMI dataset of non small cell lung cancer (nsclc) that is included with
the `InSituType` package.
```{r load_data}
library(InSituType)
library(InSituTree)

# Basic dataset
data(mini_nsclc)
str(mini_nsclc)

# Cell type reference profile
data(ioprofiles)
str(ioprofiles)
```

## Necessary inputs

Here we show the format of each of the necessary inputs to run `InSituTree`.
Many of these mirror inputs for `InSituType`, and more information can be found
in that package.

1. `x`: A matrix of counts data, cells x genes

```{r strcounts}
counts <- mini_nsclc$counts
str(counts)
counts[5:10, 10:15]
```

2. `neg`: A vector giving each cell's mean negative control value:

```{r strnegmean}
negmean <- Matrix::rowMeans(mini_nsclc$neg)
head(negmean)
```

3. `full_profiles`: Reference profile giving the expected expression profile of
each cell subtype, with genes in rows and cell subtypes in columns. Major cell
types are not needed. The reference matrix must be linear-scale, not log-scale.
Genes (rows) do not need to be aligned to the counts matrix. For more
information about generating a reference profile, see [here](https://nanostring-biostats.github.io/CosMx-Analysis-Scratch-Space/posts/hybrid-reference-profiles/).

```{r strprofiles}
str(ioprofiles)
ioprofiles[1:5, ]
```

4. `cth`: A nested list of the **c**ell **t**yping **h**ierarchy. Here we set up
one that groups the immune cells into multiple subclusters. Any single-child
nodes will be treated as endpoints. Endpoints can be named or unnamed; see "NK"
for an example of an endpoint that is unnamed.

```{r strcth}
celltype_hierarchy_list <- list(
  "structural" = c(
    "endothelial",
    "fibroblast"
  ),
  "myeloid" = c(
    "macrophage",
    "mast",
    "mDC",
    "monocyte",
    "neutrophil"
  ),
  "lymphoid" = list(
    "B-lymphoid" = c(
      "B-cell",
      "pDC",
      "plasmablast"
    ),
    "T-lymphoid" = list(
      "T4" = c(
        "T CD4 memory",
        "T CD4 naive",
        "Treg"
      ),
      "T8" = c(
        "T CD8 memory",
        "T CD8 naive"
      ),
      "NK"
    )
  )
)
```

5. `name_for_new_annotation`: A character name for the new annotations. This
will appear as the name for the top level cell typing results and is used in a
recursive function call. Recommendation is to leave at the default value of 
"topLevel".

```{r strannotNames}
annotation_names <- "topLevel"
```

## Optional inputs

The remaining inputs to `InSituTree` are optional, with some having default
values.

6. `cohort`: A vector of cohorts provides a grouping of cells based on
alternative, non-RNA, data types, such as immunofluorescent staining, cell
shape metrics, tissue structures, etc. 
One way to generate these cohorts is to use the `fastCohorting()` function in the
`InSituType` package and include any variables that might be informative for
cell typing, *except* variables you'll want to analyze later. For example, if
you'll later perform differential expression as a function of spatial context,
then it's safer to exclude spatial context from the cell typing exercise (and
therefore from this function). 
In this small example we will not use cohorts, but it is generally recommended.
The random vector below is provided to give an example of the expected
structure. See the `InSituType` package for more information about how to
generate cohorts with the `fastCohorting()` function.

```{r strcohorts}
# These will not be used in this vignette and should not be random.
# This is an example to show the structure.
set.seed(2024)
example_cohorts <- sample(1:30, size = ncol(counts), replace = TRUE)
```

7. `excluded_genes`: A character vector of genes to exclude from the cell
typing. For example, these may be genes with a strong platform effect between
the query dataset and reference profiles. Defaults to NULL.

```{r strExcludedGenes}
genes_to_exclude <- c("MALAT1")
```

8. `quantile_absolute_expression_difference_param`: A lower quantile limit for
gene selection based on absolute difference in expression between the cell type
with the highest expression and that with the lowest expression among all cell
types considered for a given branch. Value should be between 0 and 1; defaults
to 0.5.

```{r strQuantileAbsoluteDifference}
quantile_abs_difference <- 0.5

# For reference, the absolute differences in expression
# for a set of six genes across the whole ioprofile are:
diffs <- sapply(c("CD3D", "CD3E", "JCHAIN", "KRT18", "VIM", "MALAT1"),
                function(gene) {
                  diff(range(ioprofiles[gene, ]))
                })
diffs

# The 50% quantile is:
quantile(diffs, 0.5)

# Genes satisfying the expression criteria for absolute difference,
# ie.e genes with expression difference bigger than the 50% of all genes'
# expression differences, if only six were considered:
names(diffs)[diffs > quantile(diffs, quantile_abs_difference, na.rm = TRUE)]
```

9. `quantile_percent_expression_difference_param`: A lower quantile limit for
gene selection based on percentage difference in expression between the cell
type with the highest expression and that with the lowest expression among all
cell types considered for a given branch. Value should be between 0 and 1;
defaults to 0.5.

```{r strQuantilePercentageDifference}
quantile_perc_difference <- 0.5

# For reference, the percent differences in expression
# for a set of six genes across the whole ioprofile are:
percs <- sapply(c("CD3E", "JCHAIN", "KRT18", "VIM", "MALAT1"), function(gene) {
  diff(range(ioprofiles[gene, ])) / max(ioprofiles[gene, ])
})
percs

# The 50% quantile is:
quantile(percs, 0.5)

# Genes satisfying the expression criteria for percentage difference,
# ie.e genes with expression difference bigger than the 50% of all genes'
# percentage expression differences, if only six were considered:
names(percs)[percs > quantile(percs, quantile_perc_difference, na.rm = TRUE)]
```

The only gene passing **both** of the above filters is JCHAIN, and so in this
small example where we're considering all cell types together and only six
genes, only this gene would be considered as having high differentiating power
at the given granularity level. Note that when fewer cell types or subtypes are
considered simultaneously, we expect to see more variability in the diffs and
percs.

10. `return_summary_annotation`: A logical value indicating whether to return
the final summary annotation.
This data frame shows the annotation that has been made for each cell at each
level of the annotation tree. Recommendation is to keep this at its default
value of TRUE.

```{r strReturnSummary}
return_summary_df <- TRUE
```

## Tree-style cell typing

First, we can print a simple structured list to visualize the cell type
hierarchy we've laid out with our cell type hierarchy list.

```{r cellTypeStructure, fig.width=6, fig.height=4}
printTree(celltype_hierarchy_list, prefix = "|-")
```

Once we've verified that this tree shows the hierarchy and cell type
relationships that we'd expect, we can run supervised nested cell typing.

```{r runInSituTree}
res <- runInSituTree(
  x = counts,
  neg = negmean,
  full_profiles = ioprofiles,
  cth = celltype_hierarchy_list,
  name_for_new_annotation = annotation_names,
  cohort = NULL,
  excluded_genes = genes_to_exclude,
  quantile_absolute_expression_difference_param = quantile_abs_difference,
  quantile_percent_expression_difference_param = quantile_perc_difference,
  return_summary_annotation = return_summary_df
)
```

In the messages that print out, we see that a handful of cells are removed from
cell typing if they have zero counts in any of the gene subsets used for typing.

## Review results

The results from running `InSituTree` contain two main components:

1) Nested list of results from each run of `InSituTypeML`

2) Summary of annotations at each level of the tree

The nested results list contains the typical outputs of `InSituType` for each
round of cell typing, and one additional metric, `ctsPerCell`, which is the 
counts of the selected genes for each cell at the given round of subclustering.
This counts metric could be used to evaluate the confidence of a particular
round of subclustering, since cell type assignments for low-count cells are
likely to be less confident. 
Each level of the results list contains:

- `result`: The `InSituType` results plus counts per cell

- `subclusterings`: A nested list of clustering results for the next level down
of annotations.

- `name`: The name of the celltype that was subclustered.

As an example, we look at the assignment of
structural cells to fibroblast or endothelial
and use the flightpath tool from `InSituType` to visualize how well cells were
assigned to these major types.

```{r resList}
names(res)

names(res$topLevel)

str(res$topLevel$result)

str(res$topLevel$subclusterings)

InSituType::flightpath_plot(
  insitutype_result =
    res$topLevel$subclusterings$structural$result)
```

The summary annotation data frame shows the cell type assigned at each level of
annotation, as well as the posterior probability for each assignment. Any final
cell types, found at the bottom of the cell type hierarchy, are cascaded right
to further annotation levels. For example, cells assigned to 'endothelial' at
annotation level 2 in this example will also be labeled as 'endothelial' for
annotation levels 3 and 4. In other words, the the right most `annotLevel`
will contain the final cluster assignment at the finest level.

Note that any cells which did not have any counts in the genes used for
subclustering will not get a cell type assignment; if this happens in the first
round of annotation, these cells will show 'NA' for all annotation levels. If
this happens at later rounds of annotation, the last assignment will be cascaded
right.

```{r resSummary}
head(res$summaryAnnotation, 10)
```


```{r}
sessionInfo()
```
